---
title: "Setup — Posit Cloud Verification"
subtitle: "A five-minute check before we begin"
author: "R. Burke Squires"
date: today
format:
  html:
    toc: true
---

## Today we're on Posit Cloud

For the workshop itself we are using **Posit Cloud** — a browser-based RStudio environment with R, Python, and Quarto already installed. There is nothing to install on your laptop today.

::: {.callout-tip}
## Why Posit Cloud for the workshop
Setup that would take 60–90 minutes to do locally on a federal-government laptop takes about 60 seconds here. We use the time on actual training instead.
:::

For continuing on your own after the workshop, see [Next Steps — Continuing Locally](next_steps.qmd) — that's the guide for installing everything on your own Mac or PC for Monday morning.

## Step 1 — Sign in to Posit Cloud

1. Go to <https://posit.cloud>.
2. Sign in or create a free account (Google / GitHub / email all work).
3. Your instructor will share a **workshop project link** — click it to open the project in your browser.

You should see the four-pane RStudio interface (Source, Console, Environment, Files).

## Step 2 — Confirm Quarto is working

In the **Terminal** pane (next to the Console tab), type:

```bash
quarto --version
```

You should see `1.9.x` or later. If you see anything else, raise your hand — we'll sort it out.

## Step 3 — Render this file

Open this file (`00_setup.qmd`) in the Posit Cloud editor and click the **Render** button (blue arrow icon at the top of the file). Quarto will:

1. Run any code chunks.
2. Convert markdown + results to HTML.
3. Open the result in the Viewer pane.

```{r}
#| label: hello-world
#| message: false

# A trivial chunk that proves the engine is alive.
mean(1:10)
```

If you see "5.5" above this paragraph and the rendered HTML opens in the Viewer — you are ready.

## Step 4 — Confirm the R packages are installed

The afternoon uses several R packages. In the workshop Posit Cloud project they are **pre-installed** — this chunk just confirms it. Run it (click the green ▶ on the chunk, or render this file):

```{r}
#| label: check-packages
#| message: false

needed <- c("tidyverse", "knitr", "kableExtra", "DT",
            "broom", "patchwork", "plotly", "rmarkdown")

installed <- needed %in% rownames(installed.packages())
report <- data.frame(package = needed,
                     installed = ifelse(installed, "yes", "MISSING"))
print(report, row.names = FALSE)

if (all(installed)) {
  message("All packages present — you're ready.")
} else {
  missing <- needed[!installed]
  message("Missing: ", paste(missing, collapse = ", "),
          " — raise your hand.")
}
```

If every row says `yes`, you're set. If anything says `MISSING`, raise your hand — your instructor can install it in a few seconds with `install.packages("name")`.

::: {.callout-important}
## Instructor note (delete before sharing with students)
**Pre-install all eight packages into the Posit Cloud workspace template before the workshop**, so attendees never wait on `install.packages()` (it can take several minutes per package × 30 people = a stalled room). From the workspace's R console, once:

```r
install.packages(c("tidyverse", "knitr", "kableExtra", "DT",
                   "broom", "patchwork", "plotly", "rmarkdown"))
```

Then save the project as the assignment/template so every student copy inherits the installed library. If you also plan to demo PDF output, add `tinytex::install_tinytex()` as well.
:::

## Step 5 — Brief tour of the project

Look at the **Files** pane in the lower-right. You should see:

- `_quarto.yml` — the project configuration.
- `index.qmd` — the workshop landing page (this site).
- `morning_rmarkdown/` — the morning R Markdown notebooks and slides.
- `bridge/` — the R Markdown → Quarto bridge module (after the first break).
- `afternoon_quarto/` — the afternoon Quarto modules.
- `capstone.qmd` and `next_steps.qmd` — end of day.
- `references.bib` — sample bibliography for the citations module.

You don't need to understand the structure yet — Quarto Module 7 will cover it. For now, just know everything you need is in this folder.

## A note on Posit Cloud free-tier limits

The free Posit Cloud tier has a monthly compute-hour limit. For today's workshop this is more than enough — but if you keep using Posit Cloud heavily after the workshop, you may run into it. Two reasonable responses:

- Upgrade to a paid Posit Cloud plan (~$5/month at the lowest tier — check current pricing at <https://posit.cloud/plans>).
- **Move to a local install** — that's covered in [Next Steps](next_steps.qmd).

## What's next

We start with R Markdown. Open [`morning_rmarkdown/01_rmarkdown_basics.Rmd`](morning_rmarkdown/01_rmarkdown_basics.Rmd) when your instructor signals.
