---
title: "Capstone — Build It"
subtitle: "Use everything from the day on real data"
author: "YOUR NAME HERE"
date: today
format:
  html:
    toc: true
---

## What you're doing

For the next 30 minutes you will build a short, self-contained Quarto report on a real dataset, applying as many of the day's techniques as you have time for.

This is not graded. There is no "complete." The goal is to leave the workshop with a `.qmd` file **you wrote yourself**, that renders, and that you understand.

## Pick a dataset

Use whichever is closer to your day-to-day work:

1. **`biomarker_data`** — the simulated NIEHS biomonitoring set we've been using all afternoon (PFOA, PFOS, cholesterol, age, sex, site, BMI, N=200). Already loaded if you've been running the workshop notebooks; otherwise, the setup chunk below recreates it.
2. **`airquality`** — built-in R dataset, NYC ozone/temp/wind/solar for May–September 1973. No download.
3. **`palmerpenguins::penguins`** — three-species size measurements from Antarctic field stations. `install.packages("palmerpenguins")`.
4. **Your own data** — if you have a CSV that's safe to bring to a training, even better.

## The brief

Produce a short report (1–3 rendered pages) that includes:

::: {.callout-note}
## Required
- [ ] A YAML header with `title`, `author`, `date`, and `format: html` with a TOC.
- [ ] An introductory paragraph (3–5 sentences) using markdown — at least one **bold** and one *italic*.
- [ ] **One** captioned figure with a `fig-` label.
- [ ] **One** captioned table with a `tbl-` label.
- [ ] **One** inline code expression in prose (e.g., "The mean was `r mean(x)`.").
- [ ] **One** cross-reference (`@fig-...` or `@tbl-...`) in prose.
- [ ] A "Methods" section with at least one well-commented code chunk.
:::

::: {.callout-tip}
## Bonus — pick one if you have time
- [ ] Add a citation. (Sample entries in `references.bib`.)
- [ ] Render the same file to `docx` (add a `docx:` format block).
- [ ] Add a callout block (`::: {.callout-note}`).
- [ ] Add a two-column layout for a side-by-side comparison.
- [ ] Use `::: {.panel-tabset}` to show R and Python versions of the same plot.
- [ ] Add `code-fold: true` and verify the code is hidden by default but expandable.
- [ ] Re-render as a revealjs slide deck.
:::

## A starting template

```yaml
---
title: "Replace with your title"
author: "Your name"
date: today
format:
  html:
    toc: true
    code-fold: false
---
```

## Introduction

::: {.callout-warning}
## Replace this section
Write 3–5 sentences about your dataset and what you want to learn from it.
:::

## Methods

```{r}
#| label: setup
#| message: false
#| warning: false
#| include: false

library(dplyr)
library(ggplot2)
library(knitr)
```

```{r}
#| label: load-data
#| message: false

# Replace this with however you load your dataset.
# Default: recreate biomarker_data from the afternoon Quarto modules
set.seed(42)
n <- 200
dat <- data.frame(
  participant_id = 1:n,
  age = round(rnorm(n, 48, 14)),
  sex = sample(c("Female", "Male"), n, replace = TRUE, prob = c(0.55, 0.45)),
  pfoa_serum = round(rlnorm(n, meanlog = 0.8, sdlog = 0.9), 2),
  cholesterol = round(rnorm(n, 210, 38)),
  site = sample(c("Research Triangle Park", "Durham", "Chapel Hill"),
                n, replace = TRUE)
)
```

## Results

```{r}
#| label: tbl-summary
#| tbl-cap: "Summary of the dataset."

# Replace with your own summary.
summary_tbl <- dat |>
  summarise(across(where(is.numeric), \(x) round(mean(x, na.rm = TRUE), 2)))
kable(summary_tbl)
```

```{r}
#| label: fig-overview
#| fig-cap: "Overview plot of the data."
#| fig-width: 6
#| fig-height: 4
#| warning: false

# Replace with a plot relevant to your question.
ggplot(dat, aes(x = pfoa_serum, y = cholesterol)) +
  geom_point(color = "#205C40", alpha = 0.6) +
  scale_x_log10() +
  theme_minimal() +
  labs(x = "Serum PFOA (ng/mL, log scale)", y = "Total Cholesterol (mg/dL)")
```

The mean cholesterol in the dataset is `r round(mean(dat$cholesterol, na.rm = TRUE), 1)` mg/dL. See @fig-overview for the full distribution.

## Discussion

::: {.callout-warning}
## Replace this section
Two or three sentences interpreting your results.
:::

## What to do if you finish early

- Add a second figure or table.
- Try rendering to `docx` and open the result in Word.
- Convert the entire file to `revealjs` and step through it as slides.
- Add it to a Quarto project with a `_quarto.yml`.

## What to do if you get stuck

- Save the file. Comment out the chunk that just stopped working.
- Render. If that works, your problem is in the commented chunk.
- Re-add lines one at a time until you find the breaking one.
- Ask a neighbor before asking the instructor. Pair debugging is faster.

## Wrap-up

When you have something rendering that you are proud of (or at least not embarrassed by), raise your hand and we will pick a couple of volunteers to demo on the projector for the last 5 minutes.

Thanks for being here.
