---
title: "My Reproducible Report"
author: "YOUR NAME"
date: today
bibliography: references.bib
format:
  html:
    toc: true
    code-fold: false
    link-citations: true
---

```{r}
#| label: setup
#| include: false

library(dplyr)
library(ggplot2)
library(knitr)

biomarker_data <- read.csv("data/workshop_biomarkers.csv")
stopifnot(nrow(biomarker_data) == 200)
```

## Readiness check

This learner file is a separate Quarto project. Rendering it should not render the training website or any lesson page.

```{r}
#| label: readiness-table
#| echo: false

readiness <- data.frame(
  check = "Mean of the integers 1 through 10",
  result = mean(1:10)
)

kable(readiness)
```

The synthetic dataset contains `r nrow(biomarker_data)` rows. Replace this sentence with a result that matters to your report later today.

::: {.callout-note}
## Synthetic data
These records were generated solely for teaching. They contain no participant data and must not be interpreted as evidence about PFAS exposure or health outcomes.
:::

## Introduction

<!-- During Quarto 1, replace this comment with 3–5 sentences. Include one bold phrase and one italic phrase. -->

## Methods

The report uses the supplied synthetic workshop dataset. Code is executed from a clean R session when the document renders.

## Results

<!-- During Quarto 2, add a captioned table here. -->

<!-- During Quarto 3, add a captioned figure with fig-alt here. -->

<!-- During Quarto 4, add cross-references and one citation. -->

## Discussion

<!-- During the capstone, write 2–3 restrained sentences interpreting the synthetic results and noting a limitation. -->
