---
title: "From R Markdown to Quarto"
subtitle: "What changes, what doesn't"
author: "R. Burke Squires"
institute: "NIEHS"
date: today
format:
  revealjs:
    theme: [default, niehs-theme.scss]
    logo: ../images/niehs-logo.png
    footer: "Quarto Workshop · NIEHS · 2026"
    slide-number: true
    chalkboard: true
    code-overflow: wrap
---

## The headline

> If you know R Markdown, you know **85% of Quarto**.

::: {.fragment}
The remaining 15% is mostly syntax — and it's the **same** syntax across every language Quarto supports.
:::

# What stays the same {.section-divider}

## Don't relearn these

::: {.incremental}
- YAML header at the top
- Pandoc markdown for prose
- `` `r expr` `` inline code
- `` ```{r} `` chunk fences
- Same conceptual three-part anatomy
- Render button does the same thing
:::

# What changes {.section-divider}

## Change 1: `output:` → `format:`

::: {.columns}
::: {.column width="50%"}
**R Markdown**

```yaml
output:
  html_document:
    toc: true
    toc_depth: 2
    code_folding: hide
```
:::
::: {.column width="50%"}
**Quarto**

```yaml
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: true
```
:::
::::

::: {.fragment}
Top-level key, format names, underscores → hyphens.
:::

## Change 2: chunk options leave the curly braces

::: {.columns}
::: {.column width="50%"}
**R Markdown**

```` markdown
```{r myplot, fig.width=6, echo=FALSE}
plot(1:10)
```
````
:::
::: {.column width="50%"}
**Quarto**

```` markdown
```{r}
#| label: myplot
#| fig-width: 6
#| echo: false

plot(1:10)
```
````
:::
::::

## Why this change matters

::: {.incremental}
1. **Same option names across languages.** `fig-width: 6` works in `{r}` and `{python}`.
2. **Long chunk headers stop being walls of text.**
3. **Better editor support** — folding, syntax highlighting, IDE help.
:::

## Change 3: cross-references are built in

**In R Markdown:** install `bookdown`, learn its labeling conventions.

**In Quarto:** label with the right prefix, add a caption — done.

```` markdown
```{r}
#| label: fig-ozone
#| fig-cap: "Ozone vs temperature."

plot(x, y)
```

The relationship is roughly linear (@fig-ozone).
````

::: {.fragment}
This is the biggest quality-of-life win.
:::

## Change 4: file extension is `.qmd`

::: {.incremental}
- New files: `.qmd`, not `.Rmd`.
- Existing `.Rmd` files still render with `quarto render` (mostly).
- You don't have to convert everything at once.
:::

# The translation table {.smaller .section-divider}

## Chunk options

| R Markdown | Quarto |
|---|---|
| `echo=FALSE` | `#\| echo: false` |
| `eval=FALSE` | `#\| eval: false` |
| `include=FALSE` | `#\| include: false` |
| `message=FALSE` | `#\| message: false` |
| `warning=FALSE` | `#\| warning: false` |
| `fig.width=6` | `#\| fig-width: 6` |
| `fig.height=4` | `#\| fig-height: 4` |
| `fig.cap="..."` | `#\| fig-cap: "..."` |
| `cache=TRUE` | `#\| cache: true` |

`TRUE`/`FALSE` (R) → `true`/`false` (YAML).

# Migration strategies {.section-divider}

## Pick your pace

::: {.incremental}
1. **Leave old `.Rmd` alone.** They still render. Forever, basically.
2. **Convert as you touch.** Next time you edit, take 10 min.
3. **Big-bang migration.** Worth it for actively-developed projects.
:::

## A semi-automated path

::: {.incremental}
1. Copy `your_file.Rmd` → `your_file.qmd`.
2. YAML: `output: html_document` → `format: html`.
3. `quarto render your_file.qmd`.
4. Fix the errors that come up (usually a few chunk options).
:::

::: {.fragment}
~10–15 min per file, once you've done a couple.
:::

# Hands-on {.section-divider}

## Convert your morning file

::: {.callout-tip}
Open `notebooks/rm_02_rmd_to_quarto_bridge.qmd`. Do Exercise 2.1.

Goal: take the `.Rmd` you built before the break and produce a working `.qmd` version of it.
:::

This is the single most important exercise of the morning.
