---
title: "From R Markdown to Quarto"
subtitle: "What transfers, what changes"
author: "R. Burke Squires"
institute: "NIEHS"
date: today
format:
  revealjs:
    theme: [default, ../../afternoon_quarto/slides/niehs-theme.scss]
    logo: ../../images/niehs-logo.png
    footer: "R Markdown + Quarto · NIEHS · 2026"
    slide-number: true
    chalkboard: true
    code-overflow: wrap
---

## The durable model

> YAML + prose + executable code still produce the document.

::: {.fragment}
R Markdown remains supported. Quarto adds a broader publishing system; it does not invalidate existing `.Rmd` work.
:::

# What transfers {.section-divider}

## Keep these skills

::: {.incremental}
- YAML at the top
- Pandoc Markdown for prose
- Inline R results when using knitr
- `` ```{r} `` chunk fences
- Clean-session rendering
- Narrative, code, data, and output in one source
:::

# What changes {.section-divider}

## Change 1: `output:` → `format:`

::: {.columns}
::: {.column width="50%"}
**R Markdown**

```yaml
output:
  html_document:
    toc: true
    toc_depth: 2
    code_folding: hide
```
:::
::: {.column width="50%"}
**Quarto**

```yaml
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: true
```
:::
::::

## Change 2: chunk options become `#|` lines

::: {.columns}
::: {.column width="50%"}
**R Markdown**

```` markdown
```{r myplot, fig.width=6, echo=FALSE}`r ''`
plot(1:10)
```
````
:::
::: {.column width="50%"}
**Quarto**

```` markdown
```{r}`r ''`
#| label: myplot
#| fig-width: 6
#| echo: false

plot(1:10)
```
````
:::
::::

## Why the `#|` form helps

::: {.incremental}
- One option per line
- YAML-style names and values
- Consistent execution-option syntax across supported engines
- Easier-to-read chunk headers
:::

## Change 3: cross-references are built in

```` markdown
```{r}`r ''`
#| label: fig-ozone
#| fig-cap: "Ozone and temperature."
#| fig-alt: "Scatterplot of ozone against temperature."

plot(x, y)
```

See @fig-ozone.
````

::: {.fragment}
Labels need a type prefix such as `fig-` or `tbl-`.
:::

## Change 4: new Quarto sources use `.qmd`

::: {.incremental}
- Keep working `.Rmd` files when they still meet the need.
- Use `.qmd` for a new Quarto document.
- Convert a copy, render it, and compare outputs before replacing anything.
:::

# Hands-on {.section-divider}

## Convert your safe learner report

1. Knit `student_work/morning_report.Rmd`.
2. Save a copy as `student_work/bridge_converted.qmd`.
3. Translate its YAML and two chunk headers.
4. Render the `.qmd` copy.
5. Use `student_work/checkpoints/bridge_converted.qmd` for recovery.

::: {.callout-tip}
Do not edit the workshop website or its root `_quarto.yml`.
:::
