---
title: "Serum PFOA and Cholesterol in a Synthetic Workshop Dataset"
author: "YOUR NAME"
date: today
bibliography: ../references.bib
format:
  html:
    toc: true
    code-fold: true
    link-citations: true
---

```{r}
#| label: setup
#| include: false

library(dplyr)
library(ggplot2)
library(knitr)

biomarker_data <- read.csv("data/workshop_biomarkers.csv")
```

## Introduction

This report demonstrates a **reproducible workflow** using *synthetic teaching data*. It summarizes serum PFOA and cholesterol measurements for 200 simulated records from three Triangle-region sites. The goal is to practice connecting narrative, code, and output in one document—not to make a scientific claim.

::: {.callout-note}
## Synthetic data
The generated relationships are visible teaching examples and must not be interpreted as evidence about PFAS exposure or health outcomes.
:::

## Methods

The document reads a fixed CSV file and performs all calculations during rendering from a clean R session.

## Results

```{r}
#| label: tbl-site-summary
#| tbl-cap: "Summary of the synthetic biomarker data by monitoring site."

site_summary <- biomarker_data |>
  group_by(site) |>
  summarise(
    participants = n(),
    mean_pfoa_ng_ml = round(mean(pfoa_serum), 2),
    mean_cholesterol_mg_dl = round(mean(cholesterol), 1),
    .groups = "drop"
  )

kable(
  site_summary,
  col.names = c("Site", "N", "Mean PFOA (ng/mL)", "Mean cholesterol (mg/dL)")
)
```

The dataset contains `r nrow(biomarker_data)` rows, and the mean cholesterol is `r round(mean(biomarker_data$cholesterol), 1)` mg/dL.

## Discussion

The table is descriptive and is not intended for scientific inference.
