---
title: "Serum PFOA and Cholesterol in a Synthetic Workshop Dataset"
author: "YOUR NAME"
date: today
bibliography: ../references.bib
format:
  html:
    toc: true
    code-fold: true
    link-citations: true
---

```{r}
#| label: setup
#| include: false

library(dplyr)
library(ggplot2)
library(knitr)

biomarker_data <- read.csv("data/workshop_biomarkers.csv")
```

## Introduction

This report demonstrates a **reproducible workflow** using *synthetic teaching data*. It summarizes serum PFOA and cholesterol measurements for 200 simulated records from three Triangle-region sites. Reproducible computational practice helps connect claims to the code and inputs that produced them [@wilson2017good].

::: {.callout-note}
## Synthetic data
The generated relationships are visible teaching examples and must not be interpreted as evidence about PFAS exposure or health outcomes.
:::

## Methods

The document reads a fixed CSV file and performs all calculations during rendering from a clean R session.

## Results

@tbl-site-summary reports the sample size and mean values by site.

```{r}
#| label: tbl-site-summary
#| tbl-cap: "Summary of the synthetic biomarker data by monitoring site."

site_summary <- biomarker_data |>
  group_by(site) |>
  summarise(
    participants = n(),
    mean_pfoa_ng_ml = round(mean(pfoa_serum), 2),
    mean_cholesterol_mg_dl = round(mean(cholesterol), 1),
    .groups = "drop"
  )

kable(
  site_summary,
  col.names = c("Site", "N", "Mean PFOA (ng/mL)", "Mean cholesterol (mg/dL)")
)
```

The dataset contains `r nrow(biomarker_data)` rows, and the mean cholesterol is `r round(mean(biomarker_data$cholesterol), 1)` mg/dL. The point cloud and fitted line appear in @fig-pfoa-cholesterol.

```{r}
#| label: fig-pfoa-cholesterol
#| fig-cap: "Serum PFOA and total cholesterol in the synthetic workshop dataset."
#| fig-alt: "Scatterplot of serum PFOA concentration on a logarithmic horizontal axis and total cholesterol on the vertical axis. Monitoring site is shown by color and sex by point shape; a black fitted line slopes upward."
#| fig-width: 7
#| fig-height: 4.5

p <- ggplot(
  biomarker_data,
  aes(x = pfoa_serum, y = cholesterol, color = site, shape = sex)
) +
  geom_point(alpha = 0.7, size = 2) +
  geom_smooth(aes(group = 1), method = "lm", se = FALSE, color = "black") +
  scale_x_log10() +
  labs(
    x = "Serum PFOA (ng/mL, log scale)",
    y = "Total cholesterol (mg/dL)",
    color = "Site",
    shape = "Sex"
  ) +
  theme_minimal()

p
```

## Discussion

A modest positive pattern is visible because the relationship was intentionally generated for teaching. The report demonstrates automatic tables, figures, inline values, cross-references, and citations; it does not support scientific or causal conclusions.

## References
