---
title: "My Morning Report — Converted to Quarto"
author: "Your name"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: true
---

```{r}
#| label: setup
#| include: false

# Quarto cell-option booleans use lowercase in #| lines.
# Native R values inside R code still use TRUE and FALSE.
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 4
)

dat <- read.csv("data/workshop_biomarkers.csv")
n_records <- nrow(dat)
mean_age <- mean(dat$age)
median_pfoa <- median(dat$pfoa_serum)
```

## Purpose

This **reproducible report** combines *narrative text* with results calculated from a synthetic teaching dataset.

## Data preview

```{r}
#| label: data-preview
#| echo: false

knitr::kable(
  head(dat),
  caption = "First six synthetic records."
)
```

## Dynamic summary

The synthetic teaching dataset contains `r n_records` records. Mean age is `r round(mean_age, 1)` years, and the median simulated serum PFOA concentration is `r round(median_pfoa, 2)`.

## Learning goals

1. Render a complete document from a clean session.
2. Recognize the YAML and cell-option changes between R Markdown and Quarto.
3. Keep reported values synchronized with the analysis.

::: {.callout-note}
These data are synthetic and are used only to practice a reproducible workflow.
:::
