---
title: "My Morning R Markdown Report"
author: "Your name"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 4
)

dat <- read.csv("../data/workshop_biomarkers.csv")
n_records <- nrow(dat)
mean_age <- mean(dat$age)
median_pfoa <- median(dat$pfoa_serum)
```

# Purpose

This **reproducible report** combines *narrative text* with results calculated from a synthetic teaching dataset.

# Data preview

```{r data-preview, echo=FALSE}
knitr::kable(
  head(dat),
  caption = "First six synthetic records."
)
```

# Dynamic summary

The synthetic teaching dataset contains `r n_records` records. Mean age is `r round(mean_age, 1)` years, and the median simulated serum PFOA concentration is `r round(median_pfoa, 2)`.

# Learning goals

1. Render a complete document from a clean session.
2. Control which code and output readers see.
3. Keep reported values synchronized with the analysis.
