---
title: "From R Markdown to Quarto"
subtitle: "What changes, what doesn't, how to migrate"
author: "R. Burke Squires"
date: today
format:
  html:
    toc: true
    toc-depth: 3
---

## Why this module exists

You just spent the morning learning R Markdown. The rest of the day is Quarto. Before we dive in, this 25-minute bridge makes one thing explicit:

**Almost everything you just learned still works.** The conceptual model is the same. The mechanics change in three specific places, and we'll walk through each.

::: {.callout-tip}
## The headline
If you already know R Markdown, you already know 85% of Quarto. The remaining 15% is mostly syntax — and one of those changes (`#|` chunk options) is the same in every language Quarto supports, which is a big simplification once you're past the muscle-memory cost.
:::

## What's the same

Don't relearn these — they transfer directly.

| Concept | R Markdown | Quarto |
|---|---|---|
| YAML header location | Top of file, between `---` lines | Top of file, between `---` lines |
| Markdown prose | Pandoc-flavored markdown | Pandoc-flavored markdown |
| Inline code | `` `r expr` `` | `` `r expr` `` |
| Chunk fence | `` ```{r} `` ... `` ``` `` | `` ```{r} `` ... `` ``` `` |
| Setup chunk pattern | `library()` + `opts_chunk$set` | Same idea (slightly evolved) |
| Render command | `rmarkdown::render()` or the Knit button | `quarto render` or the Render button |

If you understand the three-part anatomy (YAML + prose + chunks), the conceptual move to Quarto is trivial.

## Change 1 — `output:` becomes `format:`

In R Markdown:

```yaml
---
title: "My report"
output:
  html_document:
    toc: true
    toc_depth: 2
    code_folding: hide
    fig_width: 7
---
```

In Quarto:

```yaml
---
title: "My report"
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: true
    fig-width: 7
---
```

Three small differences:

1. **`output:` → `format:`** — top-level key renamed.
2. **`html_document` → `html`** — format names shorten (no more `_document` suffix; `word_document` → `docx`, `pdf_document` → `pdf`).
3. **Underscores become hyphens** — `toc_depth` → `toc-depth`, `code_folding` → `code-fold`, `fig_width` → `fig-width`. This is consistent across every option.

::: {.callout-note}
## The "hyphens not underscores" rule
This shows up everywhere in Quarto YAML. If you find yourself with a setting that doesn't seem to work, the first thing to check is whether you accidentally used underscores from muscle memory.
:::

## Change 2 — Chunk options move out of the curly braces

This is the change you will notice most.

**R Markdown:**

```` markdown
```{r myplot, fig.width=6, fig.height=4, echo=FALSE, message=FALSE}`r ''`
library(ggplot2)
ggplot(airquality, aes(Temp, Ozone)) + geom_point()
```
````

**Quarto:**

```` markdown
```{r}`r ''`
#| label: myplot
#| fig-width: 6
#| fig-height: 4
#| echo: false
#| message: false

library(ggplot2)
ggplot(airquality, aes(Temp, Ozone)) + geom_point()
```
````

Options move from the chunk header line down into `#|` lines at the top of the chunk body. Each option gets its own line. Values use YAML conventions (`false` instead of `FALSE`, hyphens instead of dots).

### Why this change matters

It's not arbitrary syntax churn. Three benefits:

::: {.incremental}
1. **Same option names across languages.** `fig-width: 6` works in `{r}` and in `{python}`. R Markdown's `fig.width` only works in R chunks.
2. **Long chunk headers stop being a wall of text.** A real-world chunk with 8 options used to be unreadable; now each option is on its own line.
3. **Better tool support.** Editor folding, syntax highlighting, and IDE help all work better on the YAML-style format.
:::

### The translation table

| R Markdown | Quarto |
|---|---|
| `echo=FALSE` | `#\| echo: false` |
| `eval=FALSE` | `#\| eval: false` |
| `include=FALSE` | `#\| include: false` |
| `message=FALSE` | `#\| message: false` |
| `warning=FALSE` | `#\| warning: false` |
| `fig.width=6` | `#\| fig-width: 6` |
| `fig.height=4` | `#\| fig-height: 4` |
| `fig.cap="..."` | `#\| fig-cap: "..."` |
| `cache=TRUE` | `#\| cache: true` |

Note `TRUE`/`FALSE` (R style) becomes `true`/`false` (YAML style).

## Change 3 — Cross-references work without `bookdown`

In R Markdown, if you want to reference a figure from prose ("see Figure 1"), you need the `bookdown` package and special labeling. It works, but it's an extra layer.

In Quarto, cross-references are built in. Two requirements: a label with the right prefix, and a caption.

```` markdown
```{r}`r ''`
#| label: fig-ozone
#| fig-cap: "Ozone vs temperature."

plot(airquality$Temp, airquality$Ozone)
```

The relationship is roughly linear (@fig-ozone).
````

The prefix matters: `fig-` for figures, `tbl-` for tables, `eq-` for equations, `sec-` for sections, `lst-` for code listings. Without the prefix, the cross-reference doesn't work.

This is the single largest *quality-of-life* improvement when you move from R Markdown to Quarto.

## Change 4 — Quarto extension is `.qmd`, not `.Rmd`

When you create a new Quarto file, it has the `.qmd` extension instead of `.Rmd`. This is how Posit Cloud / RStudio knows which renderer to call.

::: {.callout-tip}
## You can still render `.Rmd` files with Quarto
Quarto is backwards-compatible with most R Markdown files. Run `quarto render my_file.Rmd` and it will usually work. So your existing `.Rmd` files don't break — you can keep using them, or migrate them at your own pace.
:::

## Migration strategies

You don't have to convert everything at once. Three reasonable patterns:

::: {.incremental}
1. **New work in Quarto, leave old `.Rmd` files alone.** They keep rendering with R Markdown indefinitely.
2. **Convert as you touch.** When you need to edit an old `.Rmd`, take the 10 minutes to convert it. Spread the cost over time.
3. **Big-bang migration.** Convert a whole project at once. Worth it for actively-developed work; overkill for archived analyses.
:::

### A semi-automated conversion path

Quarto can render most `.Rmd` files directly without changes. If you want to actually move to `.qmd`:

1. Copy `your_file.Rmd` to `your_file.qmd`.
2. In the YAML, change `output: html_document` to `format: html` (and other format names accordingly).
3. Run `quarto render your_file.qmd`. Read any error messages.
4. Address the errors — usually a few chunk options that need the `#|` conversion.

There is no official "automatic converter" yet, but the manual process for one file takes 10–15 minutes once you've done it a couple of times.

## When you would *not* move to Quarto

A short and honest list:

- **You have a large `.Rmd` codebase and no immediate need to change.** R Markdown is in maintenance, not deprecated. It will keep working.
- **You depend on an R Markdown extension that doesn't have a Quarto equivalent.** This is rare but happens — check the specific package's documentation.
- **You only ever produce one output format and the YAML simplicity matters more than the features.** Niche case.

For most new scientific work in 2026, Quarto is the right default.

## Exercise 2.1 — Convert your morning file

Take the `.Rmd` file you built this morning and:

1. Save a copy as `your_filename.qmd`.
2. Change `output: html_document` to `format: html` (preserving any sub-options you had, converting underscores to hyphens).
3. Convert the chunk options in **one** chunk to the `#|` format.
4. Run `quarto render your_filename.qmd` from the terminal, or click **Render** in the editor.
5. Address any errors.

This is the most important exercise of the morning. Once you've done it once, the rest of the day is "you already know how to do this — here are more features."

## What's next

We start the Quarto half of the day. Open [`../afternoon_quarto/01_first_document.qmd`](../afternoon_quarto/01_first_document.qmd) when your instructor signals.
