---
title: "From R Markdown to Quarto"
subtitle: "What transfers, what changes, and how to migrate safely"
author: "R. Burke Squires"
date: today
format:
  html:
    toc: true
    toc-depth: 3
---

## Goal for this 20-minute bridge

You just learned the three-part anatomy of R Markdown: **YAML, prose, and executable code chunks**. Quarto uses the same model. This bridge focuses on the few changes you need to recognize before continuing in `student_work/my_report.qmd`.

::: {.callout-note}
## R Markdown remains a supported choice
Existing `.Rmd` work does not need to be converted merely because Quarto exists. R Markdown remains supported. Quarto is a useful default for many new publishing projects, especially when you need built-in cross-references, multiple languages, presentations, websites, books, or manuscripts.
:::

## What transfers directly

| Concept | R Markdown | Quarto |
|---|---|---|
| YAML header | Between `---` lines at the top | Same |
| Prose | Pandoc Markdown | Same |
| Inline R result | An inline `` `r knitr::inline_expr('expr')` `` expression | Same when using knitr |
| R chunk fence | `` ```{r} `` | Same |
| Reproducibility principle | Render the complete document in a clean session | Same |
| Primary engine for R | knitr | knitr |

The durable skill is not a particular button. It is understanding how source text, code, data, and metadata combine to produce an output.

## Change 1 — `output:` becomes `format:`

**R Markdown**

```yaml
output:
  html_document:
    toc: true
    toc_depth: 2
    code_folding: hide
```

**Quarto**

```yaml
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: true
```

Three patterns to notice:

1. `output:` becomes `format:`.
2. `html_document`, `word_document`, and `pdf_document` become `html`, `docx`, and `pdf`.
3. Many option names use hyphens instead of underscores: `toc-depth`, `code-fold`, and `fig-width`.

## Change 2 — Chunk options become `#|` lines

**R Markdown**

```` markdown
```{r myplot, fig.width=6, echo=FALSE, message=FALSE}`r ''`
plot(1:10)
```
````

**Quarto**

```` markdown
```{r}`r ''`
#| label: myplot
#| fig-width: 6
#| echo: false
#| message: false

plot(1:10)
```
````

The Quarto form keeps each option on its own line and uses YAML-style values.

| R Markdown | Quarto |
|---|---|
| `echo=FALSE` | `#\| echo: false` |
| `eval=FALSE` | `#\| eval: false` |
| `include=FALSE` | `#\| include: false` |
| `message=FALSE` | `#\| message: false` |
| `warning=FALSE` | `#\| warning: false` |
| `fig.width=6` | `#\| fig-width: 6` |
| `fig.height=4` | `#\| fig-height: 4` |
| `fig.cap="..."` | `#\| fig-cap: "..."` |
| `cache=TRUE` | `#\| cache: true` |

## Change 3 — Cross-references are built in

In Quarto, give an output a label with the correct prefix and add a caption:

```` markdown
```{r}`r ''`
#| label: fig-ozone
#| fig-cap: "Ozone and temperature."
#| fig-alt: "Scatterplot of ozone against temperature."

plot(airquality$Temp, airquality$Ozone)
```

See @fig-ozone.
````

Common prefixes are `fig-`, `tbl-`, `eq-`, `sec-`, and `lst-`.

## Change 4 — New Quarto sources use `.qmd`

Use `.qmd` for a new Quarto document. Existing `.Rmd` documents can remain in R Markdown. Migrate deliberately when there is a benefit, rather than converting an archive simply for uniformity.

A practical migration pattern is:

1. Keep the original `.Rmd` file.
2. Save a copy with a `.qmd` extension.
3. Translate the YAML.
4. Translate chunk options one chunk at a time.
5. Render from a clean session and compare the output with the original.
6. Commit or archive the change only after the result has been reviewed.

## Hands-on — convert your safe learner report

Do not edit a lesson page or the root project configuration.

1. Open `student_work/morning_report.Rmd`.
2. Click **Knit** and confirm the R Markdown version works.
3. Save a copy as `student_work/bridge_converted.qmd`.
4. Change `output:` to `format:` and translate the HTML options.
5. Convert the setup and data-preview chunk options to `#|` lines.
6. Click **Render**.
7. Compare your file with `student_work/checkpoints/bridge_converted.qmd` only after you have tried the conversion.

::: {.callout-important}
## Success criterion
Both the original `.Rmd` and your converted `.qmd` render, and you can point to the YAML and chunk-option changes that made the files different.
:::

## What's next

Return to the cumulative learner report at `student_work/my_report.qmd`, then open [`../afternoon_quarto/01_first_document.qmd`](../afternoon_quarto/01_first_document.qmd) when your instructor signals.
