---
title: "Quarto 5 — Revealjs Presentations"
subtitle: "Slide decks from .qmd"
author: "R. Burke Squires"
institute: "NIEHS"
date: today
format:
  revealjs:
    theme: [default, niehs-theme.scss]
    logo: ../../images/niehs-logo.png
    footer: "R Markdown + Quarto · NIEHS · 2026"
    slide-number: true
    chalkboard: true
    code-overflow: wrap
---

## Meta-observation

The deck you are looking at right now is a `.qmd` file rendered to `revealjs`.

::: {.fragment}
Open `slides/04_presentations.qmd` and look at the source. Same syntax you already know.
:::

# The one rule {.section-divider}

## Level-2 headings start new slides

```markdown
## First slide

Content of the first slide.

## Second slide

Content of the second slide.
```

::: {.fragment}
That's it. The rest is decoration.
:::

# Section dividers {.section-divider}

## Use a single `#`

```markdown
# Part 1 — Setup
```

Creates a colored full-screen title slide.

# Incremental reveals {.section-divider}

## Wrap a list

```markdown
::: {.incremental}
- First bullet (click)
- Second bullet (click)
- Third bullet (click)
:::
```

::: {.incremental}
- Like
- These
- Bullets
:::

# Columns {.section-divider}

## Side-by-side content

```markdown
:::: {.columns}
::: {.column width="50%"}
Left.
:::
::: {.column width="50%"}
Right.
:::
::::
```

# Code highlighting {.section-divider}

## Step through code line by line

````
```{r}
#| code-line-numbers: "1|2-3|4"
library(dplyr)
aq <- read.csv("airquality.csv")
aq <- na.omit(aq)
summary(aq)
```
````

Each click advances to the next highlighted region.

## Live

```{r}
#| code-line-numbers: "1|2-3|4"
#| eval: false
library(dplyr)
aq <- read.csv("airquality.csv")
aq <- na.omit(aq)
summary(aq)
```

# Speaker notes {.section-divider}

## Invisible during talk, shown in speaker view

```markdown
::: {.notes}
Don't forget to mention the funding.
:::
```

Press `s` during the talk to open the speaker view in a second window.

::: {.notes}
Like this one. You can't see it on the main projector but you can see it on your laptop.
:::

# Useful YAML options {.section-divider}

## The shortlist

```yaml
format:
  revealjs:
    theme: simple
    slide-number: true
    chalkboard: true        # press 'b' to draw
    incremental: false      # default to non-incremental
    transition: slide
    code-overflow: wrap
    logo: niehs-logo.png
    footer: "Your footer"
```

# Per-slide modifiers {.section-divider}

## Attach a class to a heading

```markdown
## A long script {.scrollable}

## A small-text slide {.smaller}

## A section divider {.section-divider}
```

Useful when one slide needs different behavior.

# Hands-on {.section-divider}

## Open notebook 04

::: {.callout-tip}
Work through Exercises 4.1 and 4.2 in `afternoon_quarto/05_revealjs_presentations.qmd`.

Aim: convert your Module-1 file into slides.
:::

Back in 60 minutes.
