---
title: "R Markdown: Chunks and Inline Results"
subtitle: "Control execution and keep prose synchronized"
author: "R. Burke Squires"
date: "`r Sys.Date()`"
output:
  ioslides_presentation:
    widescreen: true
    css: rmd_slides_style.css
    self_contained: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  comment = "",
  fig.height = 2.8,
  fig.width = 6
)
```

## Required outcome — 25 minutes

> Modify one chunk and add one dynamic sentence in `student_work/morning_report.Rmd`.

# Code chunks {.segue}

## Anatomy

```` markdown
```{r data-preview}`r ''`
head(dat)
```
````

::: incremental
- Name chunks clearly.
- Keep names unique.
- Run individual chunks while developing.
- Knit the entire file to validate it.
:::

# Chunk options {.segue}

## Five controls to learn first

| Option | Effect |
|---|---|
| `echo=FALSE` | Hide source; show result |
| `eval=FALSE` | Show source; do not run |
| `include=FALSE` | Run; hide source and result |
| `message=FALSE` | Hide messages |
| `warning=FALSE` | Hide warnings in the output |

> Understand a warning before suppressing it.

## Setup defaults

```r
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 4
)
```

Individual chunks can override global defaults.

# Inline results {.segue}

## Source syntax

```text
The dataset contains `r knitr::inline_expr('nrow(dat)')` records.
```

The value is computed during knitting, so it stays synchronized with the data.

## Calculate once, report clearly

```r
n_records <- nrow(dat)
mean_age <- mean(dat$age)
median_pfoa <- median(dat$pfoa_serum)
```

Use rounding and units deliberately in the inline prose.

# Hands-on {.segue}

## Extend the morning report

1. Set the `data-preview` chunk to `echo=FALSE`.
2. Calculate `mean_age` in the setup chunk.
3. Add mean age to the dynamic paragraph.
4. Display the first six records with `knitr::kable()`.
5. Restart R and knit.

Recovery: `student_work/checkpoints/morning_report_complete.Rmd`.

## Boundary

All workshop biomarker data are synthetic. Generated patterns are teaching devices, not scientific findings.
