---
title: "Quarto 1 — Your First .qmd"
subtitle: "What you already know + the Quarto-specific additions"
author: "R. Burke Squires"
date: today
format:
  html:
    toc: true
    toc-depth: 3
---

## Where we are

You spent the morning learning R Markdown and the bridge module covered the three things that change:

1. `output:` → `format:`
2. Chunk options move from `{r ...}` to `#|` lines
3. File extension is `.qmd`

This 35-minute module turns the learner file you already rendered into a useful Quarto report. Keep `student_work/my_report.qmd` open and add features to that same file rather than creating another document. You will meet Quarto-specific features such as callouts, panel tabsets, and multi-format YAML.

## The Quarto authoring workflow on Posit Cloud

You already know Knit. The Quarto equivalent is **Render**:

- **File → New File → Quarto Document** opens a new `.qmd` with a starter template.
- The **Render** button (same place as the Knit button, but renamed) renders the file.
- Keyboard shortcut: `Ctrl/Cmd+Shift+K` (same as Knit — Posit Cloud's IDE uses one shortcut for both, depending on file extension).

That's it. The workflow you used this morning carries over.

## The Quarto YAML, expanded

A minimal YAML:

```yaml
---
title: "My report"
format: html
---
```

But you'll usually want more. Here's a realistic scientific report header:

```yaml
---
title: "Serum PFOA and Cholesterol in the Triangle Region"
subtitle: "Biomonitoring Analysis, 2026"
author:
  - name: Jane Smith
    affiliation: NIEHS Biostatistics Branch
  - name: John Doe
    affiliation: NIEHS Epidemiology Branch
date: today
abstract: |
  We summarize generated serum PFOA and cholesterol values
  in 200 synthetic records across three Triangle-region sites.
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: true
    fig-width: 7
    fig-align: center
  docx: default
---
```

A few things to notice:

::: {.incremental}
- **Multi-author YAML** uses a list, with each author getting `name:` and `affiliation:` keys.
- **`date: today`** is a Quarto convenience — no need for `` `r knitr::inline_expr('Sys.Date()')` `` inline code.
- **Abstract** uses YAML's `|` block-scalar so you can write multi-line prose.
- **Multiple formats** stack under `format:` — the example offers HTML and Word from one source. PDF is optional reference material because it requires an additional typesetting toolchain.
:::

### Exercise 1.1 — Build a YAML

In `student_work/my_report.qmd`, edit the YAML header so it has:

1. A `title` and `subtitle`.
2. Your name as `author`.
3. `date: today`.
4. `format: html` with `toc: true` and `code-fold: true`.

Render. Confirm you see a table of contents and a "Code" button that hides/shows code chunks.

## Callout blocks — the first Quarto-only feature

Standard R Markdown HTML documents do not provide Quarto-style callouts as a built-in authoring feature. Similar boxes can be created with HTML/CSS or extensions, but Quarto makes the source concise and portable.

Quarto has native callouts. Five types:

::: {.callout-note}
## Note
General information that's nice to know but not essential.
:::

::: {.callout-tip}
## Tip
A helpful trick that improves how you work.
:::

::: {.callout-warning}
## Warning
Something that often goes wrong — pay attention here.
:::

::: {.callout-important}
## Important
Critical information; missing this breaks something.
:::

::: {.callout-caution}
## Caution
Used sparingly for genuinely dangerous operations (data loss, irreversible changes).
:::

### Syntax

```markdown
::: {.callout-note}
## Optional title

Body content goes here. Markdown formatting works
inside callouts: **bold**, *italic*, `code`, [links](https://...).
:::
```

Open and close with `:::`. The class name (`callout-note`, `callout-tip`, etc.) determines the styling.

### Variations

```markdown
::: {.callout-warning collapse=true}
## Click to expand

This callout is collapsed by default — useful for "extra detail"
that shouldn't dominate the page.
:::

::: {.callout-tip icon=false}
## A clean callout with no icon

Set `icon=false` if the default icon clutters your layout.
:::

::: {.callout-note appearance=simple}
## A flatter look

`appearance=simple` removes the colored title bar; just the
left border remains.
:::
```

### Exercise 1.2 — Customize a callout

In `student_work/my_report.qmd`, customize the existing synthetic-data callout: add or revise its heading and add one sentence that explains why the warning matters. As a stretch goal, add a short `callout-tip`. Render and confirm the source remains readable.

## Panel tabsets — Quarto's tabbed sections

R Markdown's tabsets work but are limited to wrapping content under a heading with `{.tabset}`. Quarto's `panel-tabset` is more flexible — it works with any level of heading and renders nicely in both HTML and Word.

````markdown
::: {.panel-tabset}

## R

```{r}`r ''`
#| eval: false
library(ggplot2)
ggplot(biomarker_data, aes(pfoa_serum, cholesterol)) + geom_point()
```

## Python

```{python}`r ''`
#| eval: false
import seaborn as sns
sns.scatterplot(data=biomarker_data, x="pfoa_serum", y="cholesterol")
```

## SAS

```sas
PROC SGPLOT DATA=biomarker_data;
  SCATTER X=pfoa_serum Y=cholesterol;
RUN;
```

:::
````

The above renders as clickable R / Python / SAS tabs in HTML, and as sequential sections in Word/PDF.

::: {.callout-tip}
## When to use a tabset
Tabsets shine for side-by-side comparisons: the same analysis in two languages, methods for two cohorts, results stratified by sex. They are *not* a substitute for structuring your document — don't hide critical content behind a tab.
:::

## Quarto chunk option syntax — quick refresher

The bridge module covered this. As a reminder:

```` markdown
```{r}`r ''`
#| label: pfoa-scatter
#| fig-cap: "PFOA vs total cholesterol, by sex."
#| fig-width: 7
#| fig-height: 4
#| warning: false
#| message: false

library(ggplot2)
ggplot(biomarker_data, aes(pfoa_serum, cholesterol, color = sex)) +
  geom_point(alpha = 0.6) +
  theme_minimal()
```
````

Compare to R Markdown:

```` markdown
```{r pfoa-scatter, fig.cap="PFOA vs total cholesterol, by sex.", fig.width=7, fig.height=4, warning=FALSE, message=FALSE}`r ''`
library(ggplot2)
ggplot(biomarker_data, aes(pfoa_serum, cholesterol, color = sex)) +
  geom_point(alpha = 0.6) +
  theme_minimal()
```
````

Same effect; cleaner source. Quarto Module 2 (next) goes deeper on chunks.

## The dataset for the rest of the day

The workshop uses one fixed synthetic dataset with PFOA, PFOS, cholesterol, age, sex, site, and BMI. In the course lesson, the data are one directory above; the learner report already reads its local copy.

```{r}
#| label: define-biomarker-data
#| include: false

biomarker_data <- read.csv("../data/workshop_biomarkers.csv")
```

The chunk option `include: false` runs the code but hides both the code and its output. In `student_work/my_report.qmd`, the equivalent path is `data/workshop_biomarkers.csv`.

::: {.callout-warning}
## Synthetic teaching data
The records contain no participant data. The generated relationships exist only to make teaching examples visible and must not be interpreted as evidence about PFAS exposure or health outcomes.
:::

## What we have covered

::: {.incremental}
- The Quarto authoring workflow (Render replaces Knit)
- Expanded YAML structure (multi-author, abstract, multi-format)
- **Callouts** — Quarto-native, no HTML/CSS hacks
- **Panel tabsets** — flexible, work in multiple output formats
- The single synthetic dataset used throughout the workshop
:::

What we **didn't** cover (because the morning + bridge covered it):

- Markdown formatting (same as R Markdown)
- Basic chunk options (covered with `#|` syntax in the bridge)
- Inline code (same syntax)
- Headings, lists, links, equations basics

## What's next

Quarto Module 2 builds the required `knitr::kable()` table in your cumulative report. `kableExtra`, `DT`, and `broom` are demonstrations or bonus material after the core table works.

Open [`02_code_chunks_and_tables.qmd`](02_code_chunks_and_tables.qmd).
