---
title: "Quarto 1 — Your First .qmd"
subtitle: "What you already know + the Quarto-specific additions"
author: "R. Burke Squires"
date: today
format:
  html:
    toc: true
    toc-depth: 3
---

## Where we are

You spent the morning learning R Markdown and the bridge module covered the three things that change:

1. `output:` → `format:`
2. Chunk options move from `{r ...}` to `#|` lines
3. File extension is `.qmd`

This 30-minute module is your first **standalone Quarto document** — you'll write a `.qmd` from scratch, set up the YAML the Quarto way, and meet the features that **only exist in Quarto** (callouts, panel tabsets, multi-format YAML in one place).

## The Quarto authoring workflow on Posit Cloud

You already know Knit. The Quarto equivalent is **Render**:

- **File → New File → Quarto Document** opens a new `.qmd` with a starter template.
- The **Render** button (same place as the Knit button, but renamed) renders the file.
- Keyboard shortcut: `Ctrl/Cmd+Shift+K` (same as Knit — Posit Cloud's IDE uses one shortcut for both, depending on file extension).

That's it. The workflow you used this morning carries over.

## The Quarto YAML, expanded

A minimal YAML:

```yaml
---
title: "My report"
format: html
---
```

But you'll usually want more. Here's a realistic scientific report header:

```yaml
---
title: "Serum PFOA and Cholesterol in the Triangle Region"
subtitle: "Biomonitoring Analysis, 2024"
author:
  - name: Jane Smith
    affiliation: NIEHS Biostatistics Branch
  - name: John Doe
    affiliation: NIEHS Epidemiology Branch
date: today
abstract: |
  We characterize the association between serum PFOA
  concentration and total cholesterol in 200 participants
  across three Triangle-region sites.
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: true
    fig-width: 7
    fig-align: center
  docx: default
  pdf:
    documentclass: article
    geometry: margin=1in
---
```

A few things to notice:

::: {.incremental}
- **Multi-author YAML** uses a list, with each author getting `name:` and `affiliation:` keys.
- **`date: today`** is a Quarto convenience — no need for `` `r Sys.Date()` `` inline code.
- **Abstract** uses YAML's `|` block-scalar so you can write multi-line prose.
- **Multiple formats** stack under `format:` — you can render to HTML, Word, *and* PDF from one source. We'll cover this in Quarto Module 4.
:::

### Exercise 1.1 — Build a YAML

In a new `.qmd` file, write a YAML header with:

1. A `title` and `subtitle`.
2. Your name as `author` (with affiliation).
3. `date: today`.
4. `format: html` with `toc: true` and `code-fold: true`.

Render. Confirm you see a table of contents and a "Code" button that hides/shows code chunks.

## Callout blocks — the first Quarto-only feature

R Markdown has no native callout boxes. To create the styled "Tip" and "Warning" boxes you saw in Module 05 of the morning, the existing materials use inline HTML and CSS — workable, but ugly in source.

Quarto has native callouts. Five types:

::: {.callout-note}
## Note
General information that's nice to know but not essential.
:::

::: {.callout-tip}
## Tip
A helpful trick that improves how you work.
:::

::: {.callout-warning}
## Warning
Something that often goes wrong — pay attention here.
:::

::: {.callout-important}
## Important
Critical information; missing this breaks something.
:::

::: {.callout-caution}
## Caution
Used sparingly for genuinely dangerous operations (data loss, irreversible changes).
:::

### Syntax

```markdown
::: {.callout-note}
## Optional title

Body content goes here. Markdown formatting works
inside callouts: **bold**, *italic*, `code`, [links](https://...).
:::
```

Open and close with `:::`. The class name (`callout-note`, `callout-tip`, etc.) determines the styling.

### Variations

```markdown
::: {.callout-warning collapse=true}
## Click to expand

This callout is collapsed by default — useful for "extra detail"
that shouldn't dominate the page.
:::

::: {.callout-tip icon=false}
## A clean callout with no icon

Set `icon=false` if the default icon clutters your layout.
:::

::: {.callout-note appearance=simple}
## A flatter look

`appearance=simple` removes the colored title bar; just the
left border remains.
:::
```

### Exercise 1.2 — Replace a `<div>` callout

In your file, write a paragraph wrapped in a `::: {.callout-tip}` callout. Render. Compare what you see to the inline HTML/CSS callout in `morning_rmarkdown/05_scientific_writing.Rmd` (if you have time to look at it).

## Panel tabsets — Quarto's tabbed sections

R Markdown's tabsets work but are limited to wrapping content under a heading with `{.tabset}`. Quarto's `panel-tabset` is more flexible — it works with any level of heading and renders nicely in both HTML and Word.

````markdown
::: {.panel-tabset}

## R

```{r}`r ''`
#| eval: false
library(ggplot2)
ggplot(biomarker_data, aes(pfoa_serum, cholesterol)) + geom_point()
```

## Python

```{python}`r ''`
#| eval: false
import seaborn as sns
sns.scatterplot(data=biomarker_data, x="pfoa_serum", y="cholesterol")
```

## SAS

```sas
PROC SGPLOT DATA=biomarker_data;
  SCATTER X=pfoa_serum Y=cholesterol;
RUN;
```

:::
````

The above renders as clickable R / Python / SAS tabs in HTML, and as sequential sections in Word/PDF.

::: {.callout-tip}
## When to use a tabset
Tabsets shine for side-by-side comparisons: the same analysis in two languages, methods for two cohorts, results stratified by sex. They are *not* a substitute for structuring your document — don't hide critical content behind a tab.
:::

## Quarto chunk option syntax — quick refresher

The bridge module covered this. As a reminder:

```` markdown
```{r}`r ''`
#| label: pfoa-scatter
#| fig-cap: "PFOA vs total cholesterol, by sex."
#| fig-width: 7
#| fig-height: 4
#| warning: false
#| message: false

library(ggplot2)
ggplot(biomarker_data, aes(pfoa_serum, cholesterol, color = sex)) +
  geom_point(alpha = 0.6) +
  theme_minimal()
```
````

Compare to R Markdown:

```` markdown
```{r pfoa-scatter, fig.cap="PFOA vs total cholesterol, by sex.", fig.width=7, fig.height=4, warning=FALSE, message=FALSE}`r ''`
library(ggplot2)
ggplot(biomarker_data, aes(pfoa_serum, cholesterol, color = sex)) +
  geom_point(alpha = 0.6) +
  theme_minimal()
```
````

Same effect; cleaner source. Quarto Module 2 (next) goes deeper on chunks.

## The dataset for the rest of the day

The afternoon uses a simulated NIEHS biomonitoring dataset — `biomarker_data` — with PFOA, cholesterol, age, sex, and site information. It's the same dataset used in your existing R Markdown training (Module 04). Define it once in your file's setup chunk:

```{r}
#| label: define-biomarker-data
#| include: false

set.seed(42)
n <- 200
biomarker_data <- data.frame(
  participant_id = 1:n,
  age = round(rnorm(n, 48, 14)),
  sex = sample(c("Female", "Male"), n, replace = TRUE, prob = c(0.55, 0.45)),
  pfoa_serum = round(rlnorm(n, meanlog = 0.8, sdlog = 0.9), 2),
  cholesterol = round(rnorm(n, 210, 38)),
  site = sample(c("Research Triangle Park", "Durham", "Chapel Hill"),
                n, replace = TRUE)
)
```

The chunk option `include: false` runs the code but hides both the code and its output. Use this pattern for setup chunks throughout your Quarto documents.

## What we've covered in 30 minutes

::: {.incremental}
- The Quarto authoring workflow (Render replaces Knit)
- Expanded YAML structure (multi-author, abstract, multi-format)
- **Callouts** — Quarto-native, no HTML/CSS hacks
- **Panel tabsets** — flexible, work in multiple output formats
- The biomarker dataset we'll use for the rest of the afternoon
:::

What we **didn't** cover (because the morning + bridge covered it):

- Markdown formatting (same as R Markdown)
- Basic chunk options (covered with `#|` syntax in the bridge)
- Inline code (same syntax)
- Headings, lists, links, equations basics

## What's next

Quarto Module 2 covers code chunks in depth, including the deep dive on **tables** that your morning training had ear-marked for Module 03 — `kable`, `kableExtra`, `DT`, and `broom` for clean model output.

Open [`02_code_chunks_and_tables.qmd`](02_code_chunks_and_tables.qmd).
