---
title: "Quarto 2 — Code Chunks & Tables"
subtitle: "Deeper chunk control + publication-quality tables"
author: "R. Burke Squires"
date: today
format:
  html:
    toc: true
    toc-depth: 3
---

## Goals for this 60-minute block

**Everyone does:** add one captioned `knitr::kable()` table to `student_work/my_report.qmd` and render it successfully.

**Instructor demonstrations/bonus:** `kableExtra`, `DT`, `broom`, caching, and mixed R/Python. These are useful options, not completion requirements.

## Setup

```{r}
#| label: setup
#| message: false
#| warning: false
#| include: false

library(dplyr)
library(ggplot2)
library(knitr)

# Read the fixed synthetic workshop dataset
biomarker_data <- read.csv("../data/workshop_biomarkers.csv")
```

We use the same fixed synthetic dataset throughout the afternoon. It contains no participant data; generated relationships are teaching devices, not scientific evidence.

## The two execution engines

Quarto picks an engine automatically based on what languages your document uses:

| Engine | When it is used | Caching when enabled |
|---|---|---|
| `knitr` | `{r}` chunks; mixed R+Python uses `reticulate` | knitr cache |
| `jupyter` | Python, Julia, or other Jupyter-language documents | Jupyter cache |

Project `freeze` is a separate feature. It stores completed document execution results in a project `_freeze/` directory for reuse during project renders.

You almost never need to choose. If you want to force one:

```yaml
engine: knitr   # or: jupyter
```

::: {.callout-note}
## Mixed R + Python is a demonstration
A document with executable `{r}` and `{python}` chunks uses **knitr** with the R package **reticulate**. That requires a configured Python environment in addition to R. We show the idea, but it is not required for the learner report.
:::

## R and Python side-by-side

The afternoon's worth of examples are in R, but the syntax for Python is just as easy. Tabsets let you present both:

::: {.panel-tabset}

### R

```{r}
#| label: fig-pfoa-r
#| fig-cap: "Serum PFOA by site — R version."
#| fig-alt: "Boxplots of simulated serum PFOA concentration by monitoring site on a logarithmic vertical axis."
#| message: false

biomarker_data |>
  ggplot(aes(x = site, y = pfoa_serum, fill = site)) +
  geom_boxplot(alpha = 0.7) +
  scale_y_log10() +
  scale_fill_brewer(palette = "Set2") +
  labs(x = NULL, y = "Serum PFOA (ng/mL, log scale)") +
  theme_minimal() +
  theme(legend.position = "none")
```

### Python

```{{python}}
# This is displayed, not executed during the workshop.

import seaborn as sns
import matplotlib.pyplot as plt

# Pull the R object into Python
bio = r.biomarker_data
fig, ax = plt.subplots(figsize=(7, 4))
sns.boxplot(data=bio, x="site", y="pfoa_serum",
            palette="Set2", ax=ax)
ax.set_yscale("log")
ax.set(xlabel=None, ylabel="Serum PFOA (ng/mL, log scale)")
plt.tight_layout()
plt.show()
```

:::

The `::: {.panel-tabset}` block renders as clickable tabs in HTML and as sequential content in non-HTML formats. The Python block above is deliberately displayed without execution so the lesson does not depend on a Python environment.

## Figures: captions, cross-references, sizes

You met `fig-cap` in the bridge module. A captioned figure with the right label prefix becomes cross-referenceable.

```{r}
#| label: fig-pfoa-chol
#| fig-cap: "PFOA serum concentration and total cholesterol in the synthetic workshop data."
#| fig-alt: "Scatterplot of serum PFOA on a logarithmic horizontal axis and total cholesterol on the vertical axis, with an upward-sloping fitted line."
#| fig-width: 6
#| fig-height: 4
#| warning: false

ggplot(biomarker_data, aes(x = pfoa_serum, y = cholesterol)) +
  geom_point(alpha = 0.5, color = "#205C40") +
  geom_smooth(method = "lm", color = "#C77D34", se = FALSE) +
  scale_x_log10() +
  labs(x = "Serum PFOA (ng/mL, log scale)",
       y = "Total Cholesterol (mg/dL)") +
  theme_minimal()
```

From prose, refer to it as `@fig-pfoa-chol` → renders as "Figure 1" with a clickable link:

> The generated teaching data show a modest positive pattern after log-transformation (@fig-pfoa-chol). This is not evidence about real participants.

::: {.callout-warning}
## Label prefixes matter
For cross-references to work, labels must start with the right prefix: `fig-` for figures, `tbl-` for tables, `eq-` for equations, `sec-` for sections, `lst-` for code listings. Quarto 4 (next module) covers this in more depth.
:::

# Tables

The required path uses plain `knitr::kable()`. The remaining methods illustrate choices you can revisit after the workshop; they are not dependencies for the learner report.

## Why default R output isn't enough

```{r}
#| label: tbl-default
#| tbl-cap: "Default R output — functional but not report-quality."

# Just printing the head of the data
head(biomarker_data, 5)
```

It works. It conveys data. But it's not what you'd put in a manuscript. Let's fix that.

## Method 1 — `kable()`

`kable()` is the workhorse table function. It's built into `knitr` and works in all output formats (HTML, Word, PDF).

```{r}
#| label: tbl-kable-basic
#| tbl-cap: "Mean biomarker concentrations by site."

biomarker_data |>
  group_by(site) |>
  summarise(
    N = n(),
    `Mean PFOA (ng/mL)` = round(mean(pfoa_serum), 2),
    `Mean PFOS (ng/mL)` = round(mean(pfos_serum), 2),
    `Mean Cholesterol (mg/dL)` = round(mean(cholesterol), 1)
  ) |>
  kable(align = c("l", "r", "r", "r", "r"))
```

The `tbl-cap` chunk option supplies the numbered table caption. The `align` argument controls per-column alignment ("l" = left, "r" = right, "c" = center). The `digits` argument (not shown here) sets default decimal places for all numeric columns.

## Method 2 — `kableExtra` for publication quality

The `kableExtra` package extends `kable()` with format-specific styling. The following code is displayed but not executed in the core course build, so the website does not depend on the package.

### Styled tables

```{r}
#| label: tbl-kableextra-styled
#| eval: false
#| tbl-cap: "Same data, styled with kableExtra."

biomarker_data |>
  group_by(site) |>
  summarise(
    N = n(),
    `Mean PFOA (ng/mL)` = round(mean(pfoa_serum), 2),
    `Mean PFOS (ng/mL)` = round(mean(pfos_serum), 2),
    `Mean Cholesterol (mg/dL)` = round(mean(cholesterol), 1)
  ) |>
  kable(align = c("l", "r", "r", "r", "r")) |>
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE,
    position = "center"
  )
```

The `kable_styling()` options:

| Option | Values | Purpose |
|--------|--------|---------|
| `bootstrap_options` | `"striped"`, `"hover"`, `"condensed"`, `"bordered"` | Visual style |
| `full_width` | `TRUE` / `FALSE` | Stretch to full page width |
| `position` | `"center"`, `"left"`, `"right"`, `"float_right"` | Table alignment |
| `font_size` | Number (e.g., 12) | Text size in points |

### Conditional row formatting

Highlight rows that meet a threshold:

```{r}
#| label: tbl-conditional
#| eval: false
#| tbl-cap: "Sites with elevated PFOA shown in red."
#| warning: false

site_summary <- biomarker_data |>
  group_by(site) |>
  summarise(
    N = n(),
    mean_pfoa = round(mean(pfoa_serum), 2),
    max_pfoa = round(max(pfoa_serum), 2)
  )

site_summary |>
  kable(col.names = c("Site", "N", "Mean PFOA", "Max PFOA")) |>
  kable_styling(bootstrap_options = c("striped", "hover"),
                full_width = FALSE) |>
  row_spec(which(site_summary$max_pfoa > 15),
           bold = TRUE, color = "#DC2626") |>
  footnote(general = "Sites exceeding 15 ng/mL max PFOA shown in red.")
```

### Grouped headers

Stack headers above related columns:

```{r}
#| label: tbl-grouped-headers
#| eval: false
#| tbl-cap: "Serum PFOA and PFOS concentrations by sex, with grouped column headers."

biomarker_data |>
  group_by(sex) |>
  summarise(
    N = n(),
    pfoa_mean = round(mean(pfoa_serum), 2),
    pfoa_sd = round(sd(pfoa_serum), 2),
    pfos_mean = round(mean(pfos_serum), 2),
    pfos_sd = round(sd(pfos_serum), 2)
  ) |>
  kable(col.names = c("Sex", "N", "Mean", "SD", "Mean", "SD")) |>
  kable_styling(bootstrap_options = c("striped", "hover"),
                full_width = FALSE) |>
  add_header_above(c(" " = 2, "PFOA (ng/mL)" = 2, "PFOS (ng/mL)" = 2))
```

The `add_header_above()` function takes a named vector — names become header text, values count the columns each header spans.

## Method 3 — `DT::datatable()` for interactive HTML

For HTML appendices and exploratory data, `DT` provides searchable, sortable, pageable tables. This optional example is displayed without execution:

```{r}
#| label: tbl-dt-interactive
#| eval: false
#| tbl-cap: "Interactive participant table (searchable, sortable, paginated)."

participants_subset <- biomarker_data |>
  select(participant_id, age, sex, site, pfoa_serum, pfos_serum, cholesterol) |>
  head(50)

datatable(participants_subset,
          filter = "top",
          options = list(pageLength = 10, autoWidth = TRUE)) |>
  formatRound(columns = c("pfoa_serum", "pfos_serum"), digits = 2)
```

::: {.callout-warning}
## `DT` is HTML-only
`DT::datatable()` produces an HTML widget. Do not assume Word or PDF will create a usable screenshot automatically. For a multi-format report, generate a static table outside HTML:

```r
if (knitr::is_html_output()) {
  DT::datatable(participants_subset)
} else {
  knitr::kable(participants_subset)
}
```
:::

## Method 4 — `broom` + `kable` for clean model output

Raw `summary(model)` output is functional but not report-ready. The optional `broom` package tidies model output into a data frame. These examples are displayed without execution in the core course build.

```{r}
#| label: tbl-model-coefficients
#| eval: false
#| tbl-cap: "Linear regression: log(PFOA) ~ age + BMI + sex + site."
#| message: false

# Fit a model
model <- lm(log(pfoa_serum) ~ age + bmi + sex + site, data = biomarker_data)

# Tidy the coefficient table
tidy(model, conf.int = TRUE) |>
  mutate(across(where(is.numeric), \(x) round(x, 3))) |>
  kable(col.names = c("Term", "Estimate", "SE", "Statistic",
                       "P-value", "Lower 95%", "Upper 95%")) |>
  kable_styling(bootstrap_options = c("striped", "hover"),
                full_width = FALSE)
```

```{r}
#| label: tbl-model-fit
#| eval: false
#| tbl-cap: "Model fit statistics."

glance(model) |>
  select(r.squared, adj.r.squared, sigma, statistic, p.value, nobs) |>
  mutate(across(where(is.numeric), \(x) round(x, 3))) |>
  kable(col.names = c("R²", "Adj. R²", "σ", "F", "P-value", "N")) |>
  kable_styling(bootstrap_options = c("condensed"), full_width = FALSE)
```

`broom::tidy()` extracts the coefficient table; `broom::glance()` extracts overall model fit. Both return data frames ready for `kable()`.

## When to use which table method

| Method | Best for | Output formats |
|--------|----------|----------------|
| `kable()` | Simple, reliable | HTML, Word, PDF |
| `kableExtra` | Publication-quality styling | HTML (best), PDF |
| `DT::datatable()` | Interactive exploration, appendices | HTML only |
| `gt` package | Highly-customized publication tables | HTML, PDF |

For a first reproducible report, plain `kable()` is the reliable default. Add format-specific styling or interactivity only when it benefits the intended reader.

## Chunk options that earn their keep

Five chunk options beyond the basics you'll reach for repeatedly:

### `cache: true` — don't re-run slow code

```{r}
#| label: slow-thing
#| cache: true
#| eval: false

# A simulation that takes 30 seconds
result <- run_expensive_simulation()
```

`cache: true` uses the active execution engine's cache—knitr cache for R or Jupyter cache for Jupyter documents. Quarto project `freeze` is separate and stores completed document results in `_freeze/` for reuse during project renders.

### `code-fold: true` — collapse code by default

```{r}
#| label: foldable-chunk
#| code-fold: true
#| code-summary: "Show the code"

mean(biomarker_data$pfoa_serum)
```

Reader sees a small arrow they can click to expand the code. Excellent for "show me the science, not the code" reports.

### `output: asis` — let output write markdown

```{r}
#| label: asis-demo
#| output: asis
#| echo: false

cat("This text appears as **rendered markdown**, not as printed output.\n")
```

This is how packages like `gtsummary` produce nicely formatted output that becomes part of the document flow.

### `dev: svg` — vector figures

```{r}
#| label: svg-figure
#| dev: svg
#| fig-width: 4
#| fig-height: 3
#| eval: false

plot(biomarker_data$pfoa_serum, biomarker_data$cholesterol)
```

Vector SVG figures scale crisply at any zoom level. Useful for HTML; for Word/PDF, `dev: png` with `dpi: 300` is the standard.

### `echo: fenced` — show the chunk header too

Useful for *teaching about chunks* — that's how this very document shows you what a chunk header looks like.

## Things that go wrong (and how to fix them)

| Symptom | Likely cause | Fix |
|---|---|---|
| "object X not found" | Chunk that defines X is below the chunk that uses it | Reorder chunks |
| "duplicate chunk label" | Two chunks share a label | Rename one |
| "could not find function" | Missing `library()` call | Add it to the setup chunk |
| Figure blurry in Word/PDF | Default `dev: png` at 72 dpi | Set `dpi: 300` |
| Table looks unstyled in Word | `kableExtra` options ignored | Use plain `kable()` for Word; `kableExtra` is HTML-best |

## Learner exercise — add the required table

In `student_work/my_report.qmd`, add a chunk under **Results** that:

1. Groups the synthetic data by `site`.
2. Reports `N`, mean PFOA, and mean cholesterol.
3. Uses `knitr::kable()`.
4. Has a label beginning `tbl-` and a `tbl-cap` chunk option.
5. Does **not** manually write “Table 1” in the caption.

Render from a restarted R session. When your table works, compare it with `student_work/checkpoints/02_table_complete.qmd`.

::: {.callout-tip}
## Optional stretch
Style a copy with `kableExtra`, tidy a model with `broom`, or display an HTML-only `DT` table. These are not required for the capstone.
:::

## What's next

Quarto Module 3 — **Figures & Visualizations**: deep dive on `ggplot2`, multi-panel layouts with `patchwork`, interactive figures with `plotly`, figure sizing strategy, and saving figures for manuscript submission.

Open [`03_figures_and_visualizations.qmd`](03_figures_and_visualizations.qmd).
