---
title: "Quarto 3 — Figures & Visualizations"
subtitle: "Publication-quality plots in .qmd"
author: "R. Burke Squires"
date: today
format:
  html:
    toc: true
    toc-depth: 3
---

## Goals for this 45-minute block

**Everyone does:** add one readable `ggplot2` figure to `student_work/my_report.qmd` with a `fig-` label, caption, alt text, and clear axis labels.

**Instructor demonstrations/bonus:** `patchwork`, `plotly`, side-by-side layouts, and manuscript-export settings.

## Setup

```{r}
#| label: setup
#| message: false
#| warning: false
#| include: false

library(dplyr)
library(ggplot2)

# Read the fixed synthetic workshop dataset
biomarker_data <- read.csv("../data/workshop_biomarkers.csv")
```

::: {.callout-warning}
## Synthetic teaching data
These records contain no participant data. The generated relationships exist only to make plotting examples visible and must not be interpreted as evidence about PFAS exposure or health outcomes.
:::

## From basic to publication-quality

One of Quarto's strengths: figures are generated directly from your code. No saving PNGs and pasting into Word — the plot **is** the code.

### A basic plot

```{r}
#| label: fig-basic
#| fig-cap: "Basic ggplot scatter — functional but unfinished."
#| fig-alt: "Scatterplot of serum PFOA and total cholesterol using default ggplot styling."

ggplot(biomarker_data, aes(x = pfoa_serum, y = cholesterol)) +
  geom_point() +
  labs(title = "PFOA Serum vs Total Cholesterol",
       x = "PFOA (ng/mL)",
       y = "Total Cholesterol (mg/dL)")
```

### The same plot, publication-ready

```{r}
#| label: fig-publication
#| fig-cap: "Publication-quality version: log-scale x, sex-stratified, with regression overlay."
#| fig-alt: "Scatterplot of serum PFOA on a logarithmic horizontal axis and total cholesterol on the vertical axis. Sex is shown by both color and point shape, with fitted lines and confidence bands."
#| warning: false

ggplot(biomarker_data, aes(x = pfoa_serum, y = cholesterol, color = sex, shape = sex)) +
  geom_point(alpha = 0.6, size = 2) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 0.8) +
  scale_x_log10() +
  scale_color_manual(values = c("Female" = "#0D9488", "Male" = "#6366F1")) +
  labs(
    title = "Generated Pattern Between Serum PFOA and Total Cholesterol",
    subtitle = "Stratified by sex, N = 200",
    x = "Serum PFOA (ng/mL, log scale)",
    y = "Total Cholesterol (mg/dL)",
    color = "Sex",
    shape = "Sex",
    caption = "Synthetic teaching data"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "bottom"
  )
```

The transformations from basic to publication-ready:

::: {.incremental}
- **Log-scale x axis** — appropriate for the right-skewed PFOA distribution.
- **Sex stratification** with deliberate color choices.
- **Linear fit overlay** with confidence band.
- **Informative labels** — title, subtitle, axis units, data caption.
- **`theme_minimal`** with `base_size = 12` for readable text at reasonable sizes.
:::

## Controlling figure size and resolution

In Quarto, figure dimensions are chunk options on `#|` lines:

| Option | Purpose | Default | Example |
|--------|---------|---------|---------|
| `fig-width` | Width in inches | 7 | `fig-width: 8` |
| `fig-height` | Height in inches | 5 | `fig-height: 4` |
| `fig-align` | Alignment | `"default"` | `"center"`, `"left"`, `"right"` |
| `dpi` | Resolution | 96 | `300` for publication |
| `fig-cap` | Figure caption | — | `"My caption"` |
| `out-width` | Rendered width (CSS) | — | `"80%"`, `"500px"` |

::: {.callout-warning}
## The R Markdown → Quarto naming gotcha
R Markdown uses dot notation: `fig.width`, `fig.height`, `fig.cap`. Quarto uses hyphens: `fig-width`, `fig-height`, `fig-cap`. If you copy R Markdown chunks into Quarto and figures stop sizing right, this is almost always why.
:::

### Narrow vs wide

A narrow, tall boxplot:

```{r}
#| label: fig-narrow
#| fig-cap: "Narrow figure: PFOA by sex (fig-width 4, fig-height 6)."
#| fig-alt: "Two boxplots compare the distribution of serum PFOA for female and male simulated records on a logarithmic vertical axis."
#| fig-width: 4
#| fig-height: 6

ggplot(biomarker_data, aes(x = sex, y = pfoa_serum, fill = sex)) +
  geom_boxplot(alpha = 0.7) +
  scale_fill_manual(values = c("#0D9488", "#6366F1")) +
  scale_y_log10() +
  labs(y = "Serum PFOA (ng/mL, log scale)", x = NULL) +
  theme_minimal() +
  theme(legend.position = "none")
```

A wide, short panel of sites:

```{r}
#| label: fig-wide
#| fig-cap: "Wide figure: age vs cholesterol faceted by site (fig-width 10, fig-height 3)."
#| fig-alt: "Three side-by-side scatterplots show age and cholesterol for Chapel Hill, Durham, and Research Triangle Park, each with a smooth fitted curve."
#| fig-width: 10
#| fig-height: 3
#| warning: false

ggplot(biomarker_data, aes(x = age, y = cholesterol)) +
  geom_point(alpha = 0.4, color = "#475569") +
  geom_smooth(method = "loess", color = "#0D9488") +
  facet_wrap(~site, nrow = 1) +
  labs(x = "Age (years)", y = "Total Cholesterol (mg/dL)") +
  theme_minimal()
```

### `out-width` for display scaling

`fig-width` controls the *creation* size. `out-width` controls how big it *appears* in the rendered document:

```{r}
#| label: fig-scaled
#| fig-cap: "Created at 8×6 inches, displayed at 60% width."
#| fig-alt: "Histogram showing the right-skewed distribution of serum PFOA on a logarithmic horizontal axis."
#| fig-width: 8
#| fig-height: 6
#| out-width: "60%"

ggplot(biomarker_data, aes(x = pfoa_serum)) +
  geom_histogram(fill = "#0D9488", color = "white", bins = 30) +
  scale_x_log10() +
  labs(title = "Distribution of Serum PFOA",
       x = "PFOA (ng/mL, log scale)", y = "Count") +
  theme_minimal()
```

This pattern is useful for journals that want figures rendered at a specific aspect ratio but displayed smaller in the manuscript.

## Multi-panel figures

Two ways to combine plots in one figure: `patchwork` (for unrelated plots) and `facet_wrap` / `facet_grid` (for the same plot subset by variable).

### `patchwork` — flexible composition

This optional example is displayed without execution so the core course build does not depend on `patchwork`.

```{r}
#| label: fig-patchwork
#| eval: false
#| fig-cap: "Two-panel figure: distribution (A) and relationship (B)."
#| fig-alt: "Panel A is a histogram of serum PFOA. Panel B is a scatterplot of serum PFOA and cholesterol with a fitted line."
#| fig-width: 10
#| fig-height: 4
#| warning: false

p1 <- ggplot(biomarker_data, aes(x = pfoa_serum)) +
  geom_histogram(fill = "#0D9488", color = "white", bins = 25) +
  scale_x_log10() +
  labs(title = "A", x = "PFOA (ng/mL)", y = "Count") +
  theme_minimal()

p2 <- ggplot(biomarker_data, aes(x = pfoa_serum, y = cholesterol)) +
  geom_point(alpha = 0.4, color = "#475569") +
  geom_smooth(method = "lm", color = "#DC2626") +
  scale_x_log10() +
  labs(title = "B", x = "PFOA (ng/mL)", y = "Cholesterol (mg/dL)") +
  theme_minimal()

p1 + p2
```

The magic is `p1 + p2`. `patchwork` overloads arithmetic operators:

| Operator | Meaning |
|----------|---------|
| `p1 + p2` | Side by side |
| `p1 / p2` | Stacked vertically |
| `p1 + p2 + p3` | Three across |
| `(p1 + p2) / p3` | Two on top, one below |
| `p1 + p2 + plot_layout(ncol = 2)` | Explicit grid |
| `p1 + p2 + plot_annotation(tag_levels = "A")` | Auto-tag panels A, B, C |

### `facet_wrap` and `facet_grid` — small multiples

```{r}
#| label: fig-faceted
#| fig-cap: "Age vs cholesterol faceted by site (columns) and sex (rows)."
#| fig-alt: "Six small scatterplots show age and cholesterol by sex and monitoring site, each with a fitted line."
#| fig-width: 8
#| fig-height: 5
#| warning: false

ggplot(biomarker_data, aes(x = age, y = cholesterol)) +
  geom_point(alpha = 0.5, size = 1.5) +
  geom_smooth(method = "lm", color = "#0D9488") +
  facet_grid(sex ~ site) +
  labs(x = "Age (years)", y = "Total Cholesterol (mg/dL)") +
  theme_bw(base_size = 10)
```

`facet_wrap(~variable)` wraps panels in a single direction. `facet_grid(rowvar ~ colvar)` creates a 2D grid. Use facets when the panels show the *same plot* on subsets; use `patchwork` when the panels show *different plots*.

## Side-by-side with Quarto columns

Sometimes you want two plots side by side in your prose without bundling them into a single ggplot object. Quarto's column layout makes this easy:

````markdown
:::: {.columns}

::: {.column width="50%"}
```{r}`r ''`
#| label: fig-left
#| fig-cap: "PFOA distribution."

# plot code
```
:::

::: {.column width="50%"}
```{r}`r ''`
#| label: fig-right
#| fig-cap: "PFOS distribution."

# plot code
```
:::

::::
````

This is **Quarto-only** — R Markdown doesn't support this layout syntax natively. Each plot remains its own labeled, captioned figure, but the layout shows them side by side.

## Interactive figures with `plotly`

For HTML reports, `plotly` can turn a static ggplot into an interactive figure. This optional example is displayed without execution so the core course build does not depend on `plotly`:

```{r}
#| label: fig-plotly
#| eval: false
#| fig-cap: "Interactive scatter — hover over points to see age."
#| fig-alt: "Interactive scatterplot of serum PFOA and cholesterol; sex is shown by color and point shape, and tooltips report age and site."

p <- ggplot(biomarker_data,
            aes(x = pfoa_serum, y = cholesterol, color = sex, shape = sex,
                text = paste("Age:", age, "<br>Site:", site))) +
  geom_point(alpha = 0.6) +
  scale_x_log10() +
  scale_color_manual(values = c("#0D9488", "#6366F1")) +
  labs(x = "PFOA (ng/mL)", y = "Cholesterol (mg/dL)") +
  theme_minimal()

plotly::ggplotly(p, tooltip = c("text", "x", "y"))
```

::: {.callout-warning}
## `plotly` is HTML-only
Interactive widgets are not automatically portable to Word and PDF; they may require screenshot tooling or a separate static fallback. For multi-format reports, use a conditional pattern:

```r
if (knitr::is_html_output()) {
  ggplotly(p)
} else {
  p  # static ggplot fallback
}
```
:::

## Saving figures for manuscript submission

For external use (journal submission, presentations), use `ggsave()`:

```{r}
#| label: save-figures-demo
#| eval: false

# Save the last plot in memory
ggsave("output/figures/pfoa_cholesterol.png",
       width = 7, height = 5, dpi = 300)

# Save a specific plot object
ggsave("output/figures/panel_figure.png",
       plot = p1 + p2,
       width = 10, height = 4, dpi = 300)

# For journals that require TIFF
ggsave("output/figures/figure1.tiff",
       width = 7, height = 5, dpi = 300, compression = "lzw")

# Vector format for crisp lines at any size
ggsave("output/figures/figure1.svg",
       width = 7, height = 5)
```

::: {.callout-tip}
## Reproducibility tip
Saving manuscript figures from code in the `.qmd` source reduces the risk that submitted figure files drift away from the analysis. Test the output paths and confirm the files are regenerated during the final workflow.
:::

## Figure cross-references

We've been using labeled figures throughout. From prose, refer to them with `@fig-label`:

> The generated teaching data show a modest positive pattern (@fig-publication); related views appear in @fig-faceted.

Quarto supplies the figure numbers and clickable links automatically. Do not type figure numbers into captions because the numbering changes when figures are added or reordered.

## Learner exercise — add the required figure

In `student_work/my_report.qmd`, add a scatterplot of `pfoa_serum` and `cholesterol` under **Results**. It must have:

1. A label beginning `fig-`.
2. `fig-cap` with an informative caption that does not begin “Figure 1.”
3. `fig-alt` that describes the axes, grouping encodings, and visible pattern.
4. A logarithmic PFOA axis.
5. Clear units in the axis labels.
6. More than color alone if you encode groups; the checkpoint uses color for site and shape for sex.

Render from a restarted R session. Compare with `student_work/checkpoints/03_figure_complete.qmd` after your version works.

::: {.callout-tip}
## Optional stretch
Combine plots with `patchwork`, make an HTML-only `plotly` copy, or save a 300-dpi manuscript file. These are not required for the capstone.
:::

## What's next

Quarto Module 4 — **Equations, Cross-References, and Citations**. Equations get the deep coverage they need for scientific reports; cross-references and citations are essentially identical to R Markdown but with one major improvement (built-in figure/table cross-refs without needing `bookdown`).

Open [`04_equations_crossrefs_citations.qmd`](04_equations_crossrefs_citations.qmd).
