---
title: "Quarto 4 — Equations, Cross-References, Citations"
subtitle: "The grammar of scientific writing"
author: "R. Burke Squires"
date: today
bibliography: ../references.bib
format:
  html:
    toc: true
    toc-depth: 3
---

## Goals for this 45-minute block

**Everyone does:** add working references to the table and figure in `student_work/my_report.qmd`, then add one citation from `student_work/references.bib`.

**Instructor demonstration/bonus:** equations, section references, CSL changes, and multi-format output.

## Setup

```{r}
#| label: setup
#| message: false
#| warning: false
#| include: false

library(dplyr)
library(ggplot2)
library(knitr)

# Read the fixed synthetic workshop dataset
biomarker_data <- read.csv("../data/workshop_biomarkers.csv")
```

::: {.callout-warning}
## Synthetic teaching data
The examples use generated records and generated relationships. They are not evidence about PFAS exposure, health effects, or real participants.
:::

# Equations

Quarto renders common LaTeX-style math in HTML, PDF, and Word, although complex commands should be tested in the intended output. The syntax is identical to what you'd use in R Markdown.

## Inline equations

Surround with single dollar signs:

```markdown
The significance threshold was $\alpha = 0.05$.
```

renders as: The significance threshold was $\alpha = 0.05$.

A few common patterns:

| You write | You see |
|-----------|--------|
| `$\alpha = 0.05$` | $\alpha = 0.05$ |
| `$R^2 = 0.87$` | $R^2 = 0.87$ |
| `$\beta_1 = 0.42$` | $\beta_1 = 0.42$ |
| `$p < 0.001$` | $p < 0.001$ |
| `$\mu \pm \sigma$` | $\mu \pm \sigma$ |
| `$\text{BMI} = \frac{kg}{m^2}$` | $\text{BMI} = \frac{kg}{m^2}$ |
| `$\log_{10}(\text{PFOA})$` | $\log_{10}(\text{PFOA})$ |
| `$\chi^2_{(3)} = 12.4$` | $\chi^2_{(3)} = 12.4$ |

## Display equations

Double dollar signs render display-style (centered, larger):

```markdown
$$Y_i = \beta_0 + \beta_1 X_{i1} + \beta_2 X_{i2} + \varepsilon_i$$
```

$$Y_i = \beta_0 + \beta_1 X_{i1} + \beta_2 X_{i2} + \varepsilon_i$$

## NIEHS-relevant equations

A few formulas you might recognize from your work.

### Dose-response (Hill model)

$$\text{Response} = \frac{E_{\max} \cdot C^n}{EC_{50}^n + C^n}$$

Where $E_{\max}$ is the maximum effect, $C$ is concentration, $EC_{50}$ is the half-maximal effective concentration, and $n$ is the Hill coefficient.

### Hazard quotient

$$\text{HQ} = \frac{\text{CDI}}{\text{RfD}}$$

Where HQ is the hazard quotient, CDI is the chronic daily intake (mg/kg/day), and RfD is the reference dose (mg/kg/day).

### Lifetime Average Daily Dose (LADD)

$$\text{LADD} = \frac{C \times IR \times EF \times ED}{BW \times AT}$$

Where:

- $C$ — concentration in the exposure medium
- $IR$ — intake rate
- $EF$ — exposure frequency
- $ED$ — exposure duration
- $BW$ — body weight
- $AT$ — averaging time

### Odds ratio

$$\text{OR} = \frac{a/b}{c/d} = \frac{ad}{bc}$$

## Equation cross-references

This is a place where Quarto improves on R Markdown. Add a label after the equation:

```markdown
$$
Y_i = \beta_0 + \beta_1 X_{i1} + \varepsilon_i
$$ {#eq-linear-regression}
```

$$
Y_i = \beta_0 + \beta_1 X_{i1} + \varepsilon_i
$$ {#eq-linear-regression}

Then reference from prose: `@eq-linear-regression` renders as "Equation 1" with a hyperlink.

> The model in @eq-linear-regression assumes residuals are normally distributed with constant variance.

::: {.callout-warning}
## Label syntax for equations
The label goes in **curly braces immediately after the closing `$$`** — separated by a space:

```
$$ ... $$ {#eq-label}
```

Keep the label adjacent to the display equation and use the `eq-` prefix; then render to confirm the reference resolves.
:::

## Aligned equations

Use the LaTeX `aligned` environment for multi-line equations with alignment at `&`:

$$
\begin{aligned}
\text{logit}(P) &= \beta_0 + \beta_1 X_1 + \beta_2 X_2 \\
\text{where logit}(P) &= \log\left(\frac{P}{1-P}\right)
\end{aligned}
$$

::: {.callout-note}
## Equation rendering across formats
- **HTML** uses MathJax — full LaTeX math support, rendered in the browser.
- **PDF** (via LaTeX) — native LaTeX, every command works.
- **Word (docx)** uses Office's native math support — most common LaTeX commands work, but obscure packages do not. If a complex equation looks wrong in Word, simplify or supply an image.
:::

# Cross-references

Cross-references let you refer to figures, tables, equations, and sections from prose, with automatic numbering that updates when content moves.

## Requirements for a cross-reference to work

Two things must be present:

1. **A label with the right prefix:** `fig-` for figures, `tbl-` for tables, `eq-` for equations, `sec-` for sections, `lst-` for code listings.
2. **A caption:** `fig-cap:` for figures, `tbl-cap:` for tables; equations and sections don't need explicit captions but need the label.

### Figure cross-reference

```{r}
#| label: fig-pfoa-cholesterol
#| fig-cap: "Serum PFOA and total cholesterol in the synthetic workshop data."
#| fig-alt: "Scatterplot of serum PFOA on a logarithmic horizontal axis and total cholesterol on the vertical axis, with an upward-sloping fitted line."
#| message: false
#| warning: false

ggplot(biomarker_data, aes(x = pfoa_serum, y = cholesterol)) +
  geom_point(alpha = 0.5, color = "#205C40") +
  geom_smooth(method = "lm", color = "#C77D34", se = FALSE) +
  scale_x_log10() +
  labs(x = "Serum PFOA (ng/mL, log scale)",
       y = "Total Cholesterol (mg/dL)") +
  theme_minimal()
```

In prose: `@fig-pfoa-cholesterol` → "Figure 1" with a clickable link.

### Table cross-reference

```{r}
#| label: tbl-summary-by-sex
#| tbl-cap: "Biomarker summary by sex."

biomarker_data |>
  group_by(sex) |>
  summarise(
    N = n(),
    `Mean Age` = round(mean(age), 1),
    `Mean PFOA (ng/mL)` = round(mean(pfoa_serum), 2),
    `Mean Cholesterol (mg/dL)` = round(mean(cholesterol), 1)
  ) |>
  kable()
```

From prose: `@tbl-summary-by-sex` → "Table 1".

### Section cross-reference

Add `{#sec-name}` after a heading:

```markdown
## Methods {#sec-methods}
```

Reference as `@sec-methods` → "Section 1.2" (numbering only works when `number-sections: true` is in YAML).

### Multiple references

You can refer to several at once:

```markdown
The descriptive results appear in @fig-pfoa-cholesterol and @tbl-summary-by-sex; @eq-linear-regression shows the model form used for illustration.
```

renders cleanly with three clickable links.

## Cross-reference comparison: R Markdown vs Quarto

This is where Quarto genuinely improves the experience.

| Feature | R Markdown | Quarto |
|---------|-----------|--------|
| Figure cross-refs | Requires `bookdown` package + special chunk labels | Built in, just use `fig-` prefix |
| Table cross-refs | Same — needs `bookdown` | Built in, `tbl-` prefix |
| Equation cross-refs | `bookdown` with `(\#eq:foo)` | Built in, `{#eq-foo}` |
| Section cross-refs | Manual anchor + `#anchor-name` | `@sec-name` |
| Cross-format support | Available through R Markdown/bookdown conventions | Built-in Quarto syntax across HTML/Word/PDF |

You can do everything in R Markdown that you can in Quarto for cross-references; you just need the `bookdown` package and a slightly more verbose syntax. Quarto bakes it into the base experience.

# Citations and bibliographies

Citations work the same way in R Markdown and Quarto. The mechanics:

## Setting up

1. Create a `.bib` file with your references (`references.bib` is conventional).
2. Add `bibliography: references.bib` to your YAML.
3. Optionally specify a citation style with `csl:`. Without one, Pandoc uses its default author-date presentation.

This file already has a bibliography configured — look at the YAML at the top.

## A BibTeX entry

```bibtex
@article{wilson2017good,
  title   = {Good enough practices in scientific computing},
  author  = {Wilson, Greg and Bryan, Jennifer and others},
  journal = {PLOS Computational Biology},
  year    = {2017},
  volume  = {13},
  number  = {6},
  pages   = {e1005510},
  doi     = {10.1371/journal.pcbi.1005510}
}
```

The key (`wilson2017good`) is what you cite by. Get BibTeX entries from:

- **Zotero** — export collection → BibTeX.
- **Google Scholar** — click the quote icon under a result → BibTeX.
- **PubMed or a journal page** — export a citation in a supported format.
- **A citation manager** — import by DOI, review the metadata, then export BibTeX.

## Citing in prose

| You write | Meaning |
|---|---|
| `@nrc2009` | Narrative citation; the author becomes part of the sentence |
| `[@nrc2009]` | Parenthetical citation |
| `[@nrc2009, p. 47]` | Parenthetical citation with a locator |
| `[-@nrc2009]` | Suppress the author in an author-date style |
| `[@nrc2009; @wilson2017good]` | Cite multiple sources together |

The exact punctuation, abbreviations, and numbering depend on the active CSL style.

A few worked examples using the entries already in this project's `references.bib`:

> Quarto provides a multi-language scientific publishing system [@allaire2022quarto]. *Science and Decisions* discusses environmental risk-assessment practice [@nrc2009]. For broader reproducibility practices, see [@wilson2017good].

The bibliography is auto-generated at the end of the document. To place it somewhere specific:

```markdown
## References

::: {#refs}
:::
```

## Citation styles

Different journals require different citation formats (APA, AMA, Nature, Vancouver). Quarto uses **CSL** (Citation Style Language) files:

```yaml
csl: nature.csl
```

For reliable classroom and regulated-network use, download the required CSL file in advance and keep it with the project rather than depending on a render-time network request.

The [Zotero Style Repository](https://www.zotero.org/styles) is a common source for journal CSL files. Confirm the journal's current author instructions before submission.

::: {.callout-tip}
## Keep style files with the project
A local CSL file improves repeatability and avoids firewall or network failures. Record where it came from and update it when journal requirements change.
:::

## Manual reference entries (no `.bib`)

You can also write references inline using Pandoc's reference syntax:

```yaml
references:
- type: article-journal
  id: wilson2017good
  author:
  - family: Wilson
    given: Greg
  - family: Bryan
    given: Jennifer
  title: "Good enough practices in scientific computing"
  container-title: PLOS Computational Biology
  issued:
    year: 2017
  volume: 13
  issue: 6
  page: e1005510
  DOI: 10.1371/journal.pcbi.1005510
```

This is rarely used in practice — `.bib` files are easier — but useful for one-off documents where you don't want a separate bibliography file.

## What happens in non-HTML output

- **HTML:** citations and the bibliography use the configured citation style; links can be enabled with `link-citations: true`.
- **Word (docx):** citations are inserted as text with the configured style; bibliography is plain text at the end.
- **PDF:** citations and the bibliography are rendered through the selected PDF toolchain and citation style.

Same source, three sensible outputs.

# Brief: multiple output formats

Cross-references and citations work because Quarto rendered them — the prose `@fig-foo` becomes a numbered, linked reference regardless of output format. To produce multiple formats from one source, stack them in YAML:

```yaml
format:
  html:
    toc: true
  docx: default
  pdf:
    documentclass: article
```

Then render all at once:

```bash
quarto render my_report.qmd --to all
```

Or pick one:

```bash
quarto render my_report.qmd --to docx
```

The end-of-day projects and safe-sharing module explains standalone reports versus complete website output and format-aware fallbacks.

# Learner exercise — connect the report

In `student_work/my_report.qmd`:

1. Add a sentence before the table that refers to it with `@tbl-...`.
2. Add a sentence before or after the figure that refers to it with `@fig-...`.
3. Cite `@wilson2017good` or another key already present in `student_work/references.bib`.
4. Add a final `## References` heading.
5. Render and confirm every reference resolves. Do not type “Table 1” or “Figure 1” manually.

Compare with `student_work/checkpoints/04_crossrefs_citations_complete.qmd` after your version works.

::: {.callout-tip}
## Optional equation exercise
Add the hazard-quotient equation with `{#eq-hq}`, then refer to it with `@eq-hq`. Equations are not required for capstone completion.
:::

## What's next

Quarto Module 5 — **Revealjs Presentations**: turn `.qmd` into slide decks.

Open [`05_revealjs_presentations.qmd`](05_revealjs_presentations.qmd).
