---
title: "Quarto 4 — Equations, Cross-References, Citations"
subtitle: "The grammar of scientific writing"
author: "R. Burke Squires"
date: today
bibliography: ../references.bib
format:
  html:
    toc: true
    toc-depth: 3
---

## Goals for this 45-minute block

- Write LaTeX equations (inline and display) including NIEHS-relevant formulas.
- Cross-reference equations, figures, tables, and sections from prose.
- Manage a bibliography with BibTeX and CSL citation styles.
- Understand which features survive format changes (HTML, Word, PDF).

## Setup

```{r}
#| label: setup
#| message: false
#| warning: false
#| include: false

library(tidyverse)
library(knitr)
library(kableExtra)

# Workshop dataset
set.seed(42)
n <- 200
biomarker_data <- data.frame(
  participant_id = 1:n,
  age = round(rnorm(n, 48, 14)),
  sex = sample(c("Female", "Male"), n, replace = TRUE, prob = c(0.55, 0.45)),
  pfoa_serum = round(rlnorm(n, meanlog = 0.8, sdlog = 0.9), 2),
  cholesterol = round(rnorm(n, 210, 38)),
  site = sample(c("Research Triangle Park", "Durham", "Chapel Hill"),
                n, replace = TRUE)
)
```

# Equations

Quarto renders LaTeX math beautifully in HTML and PDF, and reasonably in Word. The syntax is identical to what you'd use in R Markdown.

## Inline equations

Surround with single dollar signs:

```markdown
The significance threshold was $\alpha = 0.05$.
```

renders as: The significance threshold was $\alpha = 0.05$.

A few common patterns:

| You write | You see |
|-----------|--------|
| `$\alpha = 0.05$` | $\alpha = 0.05$ |
| `$R^2 = 0.87$` | $R^2 = 0.87$ |
| `$\beta_1 = 0.42$` | $\beta_1 = 0.42$ |
| `$p < 0.001$` | $p < 0.001$ |
| `$\mu \pm \sigma$` | $\mu \pm \sigma$ |
| `$\text{BMI} = \frac{kg}{m^2}$` | $\text{BMI} = \frac{kg}{m^2}$ |
| `$\log_{10}(\text{PFOA})$` | $\log_{10}(\text{PFOA})$ |
| `$\chi^2_{(3)} = 12.4$` | $\chi^2_{(3)} = 12.4$ |

## Display equations

Double dollar signs render display-style (centered, larger):

```markdown
$$Y_i = \beta_0 + \beta_1 X_{i1} + \beta_2 X_{i2} + \varepsilon_i$$
```

$$Y_i = \beta_0 + \beta_1 X_{i1} + \beta_2 X_{i2} + \varepsilon_i$$

## NIEHS-relevant equations

A few formulas you might recognize from your work.

### Dose-response (Hill model)

$$\text{Response} = \frac{E_{\max} \cdot C^n}{EC_{50}^n + C^n}$$

Where $E_{\max}$ is the maximum effect, $C$ is concentration, $EC_{50}$ is the half-maximal effective concentration, and $n$ is the Hill coefficient.

### Hazard quotient

$$\text{HQ} = \frac{\text{CDI}}{\text{RfD}}$$

Where HQ is the hazard quotient, CDI is the chronic daily intake (mg/kg/day), and RfD is the reference dose (mg/kg/day).

### Lifetime Average Daily Dose (LADD)

$$\text{LADD} = \frac{C \times IR \times EF \times ED}{BW \times AT}$$

Where:

- $C$ — concentration in the exposure medium
- $IR$ — intake rate
- $EF$ — exposure frequency
- $ED$ — exposure duration
- $BW$ — body weight
- $AT$ — averaging time

### Odds ratio

$$\text{OR} = \frac{a/b}{c/d} = \frac{ad}{bc}$$

## Equation cross-references

This is a place where Quarto improves on R Markdown. Add a label after the equation:

```markdown
$$
Y_i = \beta_0 + \beta_1 X_{i1} + \varepsilon_i
$$ {#eq-linear-regression}
```

$$
Y_i = \beta_0 + \beta_1 X_{i1} + \varepsilon_i
$$ {#eq-linear-regression}

Then reference from prose: `@eq-linear-regression` renders as "Equation 1" with a hyperlink.

> The model in @eq-linear-regression assumes residuals are normally distributed with constant variance.

::: {.callout-warning}
## Label syntax for equations
The label goes in **curly braces immediately after the closing `$$`** — separated by a space:

```
$$ ... $$ {#eq-label}
```

A common mistake is putting the label on its own line, which doesn't work.
:::

## Aligned equations

Use the LaTeX `aligned` environment for multi-line equations with alignment at `&`:

$$
\begin{aligned}
\text{logit}(P) &= \beta_0 + \beta_1 X_1 + \beta_2 X_2 \\
\text{where logit}(P) &= \log\left(\frac{P}{1-P}\right)
\end{aligned}
$$

::: {.callout-note}
## Equation rendering across formats
- **HTML** uses MathJax — full LaTeX math support, rendered in the browser.
- **PDF** (via LaTeX) — native LaTeX, every command works.
- **Word (docx)** uses Office's native math support — most common LaTeX commands work, but obscure packages do not. If a complex equation looks wrong in Word, simplify or supply an image.
:::

# Cross-references

Cross-references let you refer to figures, tables, equations, and sections from prose, with automatic numbering that updates when content moves.

## Requirements for a cross-reference to work

Two things must be present:

1. **A label with the right prefix:** `fig-` for figures, `tbl-` for tables, `eq-` for equations, `sec-` for sections, `lst-` for code listings.
2. **A caption:** `fig-cap:` for figures, `tbl-cap:` for tables; equations and sections don't need explicit captions but need the label.

### Figure cross-reference

```{r}
#| label: fig-pfoa-cholesterol
#| fig-cap: "Serum PFOA and total cholesterol are positively associated."
#| message: false
#| warning: false

ggplot(biomarker_data, aes(x = pfoa_serum, y = cholesterol)) +
  geom_point(alpha = 0.5, color = "#205C40") +
  geom_smooth(method = "lm", color = "#C77D34", se = FALSE) +
  scale_x_log10() +
  labs(x = "Serum PFOA (ng/mL, log scale)",
       y = "Total Cholesterol (mg/dL)") +
  theme_minimal()
```

In prose: `@fig-pfoa-cholesterol` → "Figure 1" with a clickable link.

### Table cross-reference

```{r}
#| label: tbl-summary-by-sex
#| tbl-cap: "Biomarker summary by sex."

biomarker_data |>
  group_by(sex) |>
  summarise(
    N = n(),
    `Mean Age` = round(mean(age), 1),
    `Mean PFOA (ng/mL)` = round(mean(pfoa_serum), 2),
    `Mean Cholesterol (mg/dL)` = round(mean(cholesterol), 1)
  ) |>
  kable() |>
  kable_styling(bootstrap_options = c("striped"), full_width = FALSE)
```

From prose: `@tbl-summary-by-sex` → "Table 1".

### Section cross-reference

Add `{#sec-name}` after a heading:

```markdown
## Methods {#sec-methods}
```

Reference as `@sec-methods` → "Section 1.2" (numbering only works when `number-sections: true` is in YAML).

### Multiple references

You can refer to several at once:

```markdown
The results (@fig-pfoa-cholesterol, @tbl-summary-by-sex) support
the hypothesis stated in @eq-linear-regression.
```

renders cleanly with three clickable links.

## Cross-reference comparison: R Markdown vs Quarto

This is where Quarto genuinely improves the experience.

| Feature | R Markdown | Quarto |
|---------|-----------|--------|
| Figure cross-refs | Requires `bookdown` package + special chunk labels | Built in, just use `fig-` prefix |
| Table cross-refs | Same — needs `bookdown` | Built in, `tbl-` prefix |
| Equation cross-refs | `bookdown` with `(\#eq:foo)` | Built in, `{#eq-foo}` |
| Section cross-refs | Manual anchor + `#anchor-name` | `@sec-name` |
| Cross-format support | Inconsistent | Same syntax across HTML/Word/PDF |

You can do everything in R Markdown that you can in Quarto for cross-references; you just need the `bookdown` package and a slightly more verbose syntax. Quarto bakes it into the base experience.

# Citations and bibliographies

Citations work the same way in R Markdown and Quarto. The mechanics:

## Setting up

1. Create a `.bib` file with your references (`references.bib` is conventional).
2. Add `bibliography: references.bib` to your YAML.
3. Specify a citation style with `csl:` (optional; defaults to Chicago author-date).

This file already has a bibliography configured — look at the YAML at the top.

## A BibTeX entry

```bibtex
@article{smith2023pfas,
  title = {PFAS Exposure and Health Outcomes in the Triangle Region},
  author = {Smith, Jane and Doe, John},
  journal = {Environmental Health Perspectives},
  year = {2023},
  volume = {131},
  pages = {1--12},
  doi = {10.1289/EHP12345}
}
```

The key (`smith2023pfas`) is what you cite by. Get BibTeX entries from:

- **Zotero** — export collection → BibTeX.
- **Google Scholar** — click the quote icon under a result → BibTeX.
- **PubMed** — use a free converter like `doi2bib.org`.
- **Crossref** — many citation managers can pull from DOI directly.

## Citing in prose

| You write | You get (author-year style) |
|-----------|----------------------------|
| `@nrc2009` | NRC (2009) |
| `[@nrc2009]` | (NRC 2009) |
| `[@nrc2009, p. 47]` | (NRC 2009, p. 47) |
| `[-@nrc2009]` | (2009) |
| `[@nrc2009; @wilson2017good]` | (NRC 2009; Wilson et al. 2017) |

A few worked examples using the entries already in this project's `references.bib`:

> Quarto is the successor to R Markdown for multi-language scientific publishing [@allaire2022quarto]. The framework for environmental risk assessment was laid out authoritatively in *Science and Decisions* [@nrc2009]. For broader reproducibility practices, see [@wilson2017good].

The bibliography is auto-generated at the end of the document. To place it somewhere specific:

```markdown
## References

::: {#refs}
:::
```

## Citation styles

Different journals require different citation formats (APA, AMA, Nature, Vancouver). Quarto uses **CSL** (Citation Style Language) files:

```yaml
csl: nature.csl
# or load directly from the Zotero repository:
csl: https://www.zotero.org/styles/jama
```

The [Zotero Style Repository](https://www.zotero.org/styles) has 10,000+ journal styles. Find your target journal, copy the URL.

::: {.callout-tip}
## CSL files don't need to be in your project
You can reference a CSL file by URL directly. Quarto downloads and caches it. Saves you from having a `nature.csl` file cluttering every project.
:::

## Manual reference entries (no `.bib`)

You can also write references inline using Pandoc's reference syntax:

```yaml
references:
- type: article-journal
  id: smith2023pfas
  author:
  - family: Smith
    given: Jane
  - family: Doe
    given: John
  title: "PFAS Exposure and Health Outcomes"
  container-title: Environmental Health Perspectives
  issued:
    year: 2023
  volume: 131
  page: 1-12
  DOI: 10.1289/EHP12345
```

This is rarely used in practice — `.bib` files are easier — but useful for one-off documents where you don't want a separate bibliography file.

## What happens in non-HTML output

- **HTML:** citations are hyperlinks; bibliography auto-numbered at the end.
- **Word (docx):** citations are inserted as text with the configured style; bibliography is plain text at the end.
- **PDF (via LaTeX):** citations and bibliography are typeset via biblatex/natbib.

Same source, three sensible outputs.

# Brief: multiple output formats

Cross-references and citations work because Quarto rendered them — the prose `@fig-foo` becomes a numbered, linked reference regardless of output format. To produce multiple formats from one source, stack them in YAML:

```yaml
format:
  html:
    toc: true
  docx: default
  pdf:
    documentclass: article
```

Then render all at once:

```bash
quarto render my_report.qmd --to all
```

Or pick one:

```bash
quarto render my_report.qmd --to docx
```

Quarto Module 7 covers the format-specific deep dive (themes, CSS, conditional output, when each format shines).

# Exercises

### Exercise 4.1 — Equation with cross-reference

Add the hazard quotient equation to your file with the label `{#eq-hq}`. In a paragraph above it, reference it: "We assess risk using the hazard quotient `(@eq-hq)`..."

Render and confirm the cross-reference is clickable.

### Exercise 4.2 — Multiple cross-references

In your file, write a paragraph that references at least one figure, one table, and one equation. Use the `(@fig-..., @tbl-..., @eq-...)` pattern. Render. Confirm all three are clickable.

### Exercise 4.3 — Add a citation

1. Open `references.bib` in the project root.
2. Add a new BibTeX entry — any paper of your choosing.
3. Cite it in your document using `[@yourkey]`.
4. Render and confirm it appears in the reference list at the end.

## Optional bonus — quartobot

There's a community tool ([quartobot by Sean Davis](https://github.com/seandavi/quartobot)) that lets you write `@doi:10.21105/joss.01686` directly in your prose, with no hand-curated BibTeX entry needed. It's marked optional for this workshop. See [`quartobot_bonus.qmd`](quartobot_bonus.qmd) for details.

## What's next

Quarto Module 5 — **Revealjs Presentations**: turn `.qmd` into slide decks.

Open [`05_revealjs_presentations.qmd`](05_revealjs_presentations.qmd).
