---
title: "Quarto 6 — Parameterized Reports"
subtitle: "Generate many reports from one source"
author: "R. Burke Squires"
date: today
format:
  html:
    toc: true
    toc-depth: 3
params:
  chemical: "PFOA"
  threshold: 4.0
  site: "Research Triangle Park"
---

## Goals for this 25-minute demonstration

- Understand what a parameterized report is and when it is useful.
- See parameters defined in YAML and used in prose, code, and captions.
- Change one parameter value and render once.
- Recognize batch generation as a future automation pattern, not a required exercise.

## The problem parameterization solves

You've written a report on **PFOA** levels at **Research Triangle Park**. Now you need to write the same report for:

- PFOS at the same site
- PFOA at Durham
- PFOA at Chapel Hill
- PFOS at all three sites

These reports have the same structure but different inputs. Without parameters, near-duplicate files drift apart. With parameters, the analysis logic remains in one source file.

Parameterized reports are especially useful for recurring reports across sites, sub-populations, chemicals, scenarios, or time periods.

## Defining parameters

Parameters live in the YAML header under a `params:` key:

```yaml
---
title: "Biomonitoring Report"
format: html
params:
  chemical: "PFOA"
  threshold: 4.0
  site: "Research Triangle Park"
---
```

Look at the YAML at the top of **this** file — it has exactly these three parameters.

::: {.callout-tip}
## YAML types matter
- `"PFOA"` (with quotes) is a string.
- `4.0` (no quotes, decimal) is a number.
- `true` / `false` (no quotes, lowercase) are booleans.
- `null` is null.

Get the type wrong and you'll get cryptic errors in your code. When in doubt, quote it as a string and `as.numeric()` it in R.
:::

## Using parameters in code and prose

Parameters appear in R as `params$chemical`, `params$site`, and `params$threshold`.

### In prose with inline code

```markdown
This report analyzes **`r knitr::inline_expr('params$chemical')`** data from
**`r knitr::inline_expr('params$site')`** using a threshold of
**`r knitr::inline_expr('params$threshold')` ng/mL**.
```

renders as:

This report analyzes **`r params$chemical`** data from **`r params$site`** using a threshold of **`r params$threshold` ng/mL**.

### In chunks

```{r}
#| label: setup
#| message: false
#| include: false

library(dplyr)
library(ggplot2)

biomarker_data <- read.csv("../data/workshop_biomarkers.csv")

chemical_columns <- c(PFOA = "pfoa_serum", PFOS = "pfos_serum")
if (!params$chemical %in% names(chemical_columns)) {
  stop("For this demonstration, chemical must be PFOA or PFOS.")
}

valid_sites <- c("All", sort(unique(biomarker_data$site)))
if (!params$site %in% valid_sites) {
  stop("For this demonstration, site must be All or one of: ",
       paste(valid_sites[-1], collapse = ", "))
}
if (!is.numeric(params$threshold) || length(params$threshold) != 1 ||
    is.na(params$threshold) || params$threshold <= 0) {
  stop("For this demonstration, threshold must be one positive number.")
}

filtered_data <- biomarker_data |>
  filter(params$site == "All" | site == params$site) |>
  transmute(
    participant_id,
    site,
    concentration = .data[[chemical_columns[[params$chemical]]]]
  )

n_above <- sum(filtered_data$concentration > params$threshold)
pct_above <- round(n_above / nrow(filtered_data) * 100, 1)
median_conc <- round(median(filtered_data$concentration), 2)
```

::: {.callout-warning}
## Synthetic data and teaching threshold
The records are synthetic, and `r params$threshold` ng/mL is an arbitrary demonstration value—not a regulatory, clinical, or health-based threshold.
:::

## Results from this parameter set

Of the `r nrow(filtered_data)` simulated records at **`r params$site`**, **`r n_above` (`r pct_above`%)** exceeded the teaching threshold for `r params$chemical`. The median concentration was `r median_conc` ng/mL.

```{r}
#| label: fig-distribution
#| fig-cap: !expr paste("Distribution of serum", params$chemical, "at", params$site)
#| fig-width: 7
#| fig-height: 4
#| fig-alt: !expr paste("Histogram of synthetic serum", params$chemical, "concentrations at", params$site, "with a dashed vertical teaching threshold at", params$threshold, "ng/mL.")
#| echo: false

ggplot(filtered_data, aes(x = concentration)) +
  geom_histogram(fill = "#0D9488", color = "white", bins = 25) +
  geom_vline(xintercept = params$threshold, color = "#DC2626",
             linewidth = 1, linetype = "dashed") +
  annotate("text", x = params$threshold + 0.3, y = Inf,
           label = paste("Threshold:", params$threshold, "ng/mL"),
           vjust = 2, hjust = 0, color = "#DC2626", size = 3.5) +
  labs(title = paste(params$chemical, "Serum Concentrations,",
                     params$site),
       x = paste(params$chemical, "(ng/mL)"),
       y = "Count") +
  theme_minimal()
```

::: {.callout-note}
## Dynamic captions
The `fig-cap` above uses `!expr` to evaluate an R expression: `!expr paste("Distribution of", params$chemical, "...")`. This is Quarto's way to compute caption text from parameters at render time. Without this, the caption would be a static string.
:::

## Re-rendering with different parameters

### From the R console

```{r}
#| eval: false

# Render for PFOS at Durham
quarto::quarto_render(
  "afternoon_quarto/06_parameterized_reports.qmd",
  execute_params = list(
    chemical = "PFOS",
    threshold = 6.0,
    site = "Durham"
  ),
  output_file = "report_PFOS_Durham.html"
)
```

Each call produces a different rendered output from the same source.

### From the command line

```bash
quarto render afternoon_quarto/06_parameterized_reports.qmd -P chemical:PFOS -P threshold:6.0 -P site:Durham
```

The `-P` flag sets a parameter. Repeat it for each value you want to override.

### IDE parameter dialog

Some RStudio/Posit environments provide **Render with Parameters** in the Render menu. Use it when available for a one-off render; the command line and R interfaces remain the reproducible options to record in automation.

## Batch report generation

The real power: a loop that generates a report for every combination you care about.

```{r}
#| label: batch-demo
#| eval: false

# Generate a separate HTML report for each chemical-site combination
chemicals <- c("PFOA", "PFOS")
sites <- c("Research Triangle Park", "Durham", "Chapel Hill")

# Arbitrary teaching thresholds for this synthetic example
thresholds <- list(PFOA = 4.0, PFOS = 6.0)

for (chem in chemicals) {
  for (site in sites) {
    out_file <- paste0("report_", chem, "_",
                       gsub(" ", "_", site), ".html")

    quarto::quarto_render(
      "afternoon_quarto/06_parameterized_reports.qmd",
      execute_params = list(
        chemical = chem,
        site = site,
        threshold = thresholds[[chem]]
      ),
      output_file = out_file
    )
  }
}
# Result: 6 HTML reports, one per chemical-site combination
```

Six reports from one source file. The important idea is that the analysis logic stays in one place.

::: {.callout-tip}
## When this pattern shines
Parameterized reports earn their keep when:

- You have a **template** that should apply to many cases.
- The **analysis is identical**; only inputs change.
- Reports need to be **regenerated regularly** as new data arrives.

NIEHS-relevant cases: annual chemical-specific exposure summaries, multi-site monitoring reports, sub-population analyses, sensitivity analyses across multiple thresholds.
:::

## Parameter validation

Parameters from a user can be wrong. A robust template validates them at the top:

```{r}
#| eval: false

# At the top of your setup chunk
valid_chemicals <- c("PFOA", "PFOS")
if (!params$chemical %in% valid_chemicals) {
  stop("Invalid chemical: ", params$chemical,
       ". Must be one of: ", paste(valid_chemicals, collapse = ", "))
}

if (!is.numeric(params$threshold) || params$threshold <= 0) {
  stop("Invalid threshold: ", params$threshold,
       ". Must be a positive number.")
}
```

A failed render with a clear error is better than a successful render of nonsense.

## R Markdown vs Quarto for parameters

Same syntax, almost identical behavior:

| Aspect | R Markdown | Quarto |
|--------|-----------|--------|
| YAML key | `params:` | `params:` (same) |
| Access in R | `params$x` | `params$x` (same) |
| Render with params (R) | `rmarkdown::render(..., params = list(...))` | `quarto::quarto_render(..., execute_params = list(...))` |
| Render with params (CLI) | N/A in core R Markdown | `quarto render ... -P key:value` |
| "Render with Parameters" button | Yes | Yes |
| Dynamic figure captions | Use `fig.cap = paste(...)` chunk option | `#\| fig-cap: !expr paste(...)` |

Both R Markdown and Quarto support parameters. Quarto also provides a convenient command-line parameter interface, which can simplify scripted report generation.

## Learner edit — change one value

Open `student_work/parameter_demo.qmd`. Change **one** YAML parameter—`chemical`, `threshold`, or `site`—and render once. Observe which prose, inline values, caption, axis label, and threshold line change.

This learner copy is safe to edit. The course lesson and website configuration remain read-only. Batch loops and programmatic rendering are reference material for later use.

## What's next

Next is the cumulative [capstone](../capstone.qmd). After the capstone, Module 7 covers projects and safe sharing.
