---
title: "Quarto 6 — Parameterized Reports"
subtitle: "Generate many reports from one source"
author: "R. Burke Squires"
date: today
format:
  html:
    toc: true
    toc-depth: 3
params:
  chemical: "PFOA"
  threshold: 4.0
  study_year: 2024
  site: "Research Triangle Park"
---

## Goals for this 25-minute block

- Understand what a parameterized report is and when to use one.
- Define parameters in YAML and use them throughout a document.
- Render the same `.qmd` file with different parameter values from the command line and from R.
- Use parameterization for batch report generation across chemicals, sites, or years.

## The problem parameterization solves

You've written a report on **PFOA** levels at **Research Triangle Park** in **2024**. Now you need to write the same report for:

- PFOS at the same site
- PFOA at Durham
- PFOA at Chapel Hill
- PFOS at all three sites
- PFOA in 2023 (for comparison)

That's six reports with structurally identical analysis but different inputs. Without parameterization, you have six near-duplicate `.qmd` files to maintain. With parameterization, you have **one** `.qmd` file and a parameter list.

This is the single most underused R Markdown / Quarto feature for environmental health research. If you produce repeated reports across chemicals, sites, sub-populations, or time periods, parameterized reports save weeks per year.

## Defining parameters

Parameters live in the YAML header under a `params:` key:

```yaml
---
title: "Biomonitoring Report"
format: html
params:
  chemical: "PFOA"
  threshold: 4.0
  study_year: 2024
  site: "Research Triangle Park"
---
```

Look at the YAML at the top of **this** file — it has exactly these four parameters.

::: {.callout-tip}
## YAML types matter
- `"PFOA"` (with quotes) is a string.
- `4.0` (no quotes, decimal) is a number.
- `2024` (no quotes, integer) is a number.
- `true` / `false` (no quotes, lowercase) are booleans.
- `null` is null.

Get the type wrong and you'll get cryptic errors in your code. When in doubt, quote it as a string and `as.numeric()` it in R.
:::

## Using parameters in code and prose

Parameters appear in R as `params$chemical`, `params$threshold`, etc.

### In prose with inline code

```markdown
This report analyzes **`r params$chemical`** data from
**`r params$study_year`** at **`r params$site`** using a threshold of
**`r params$threshold` ng/mL**.
```

renders as:

This report analyzes **`r params$chemical`** data from **`r params$study_year`** at **`r params$site`** using a threshold of **`r params$threshold` ng/mL**.

### In chunks

```{r}
#| label: setup
#| message: false
#| include: false

library(tidyverse)

# Simulate data based on the chemical parameter
# (Real workflow would read from a file; this is for demo)
set.seed(42)
n <- 200

# Different chemicals have different distributional parameters
chem_params <- list(
  PFOA = list(meanlog = 0.8, sdlog = 0.9),
  PFOS = list(meanlog = 1.1, sdlog = 0.7),
  PFNA = list(meanlog = 0.5, sdlog = 0.6),
  BPA  = list(meanlog = 0.3, sdlog = 1.1)
)

if (params$chemical %in% names(chem_params)) {
  meanlog <- chem_params[[params$chemical]]$meanlog
  sdlog <- chem_params[[params$chemical]]$sdlog
} else {
  meanlog <- 0.5
  sdlog <- 0.8
}

exposure_data <- data.frame(
  participant_id = 1:n,
  site = sample(c("Research Triangle Park", "Durham", "Chapel Hill"),
                n, replace = TRUE),
  concentration = round(rlnorm(n, meanlog = meanlog, sdlog = sdlog), 2)
)

# Filter to the site of interest if specified
if (!is.null(params$site) && params$site != "All") {
  filtered_data <- exposure_data |>
    filter(site == params$site)
} else {
  filtered_data <- exposure_data
}

n_above <- sum(filtered_data$concentration > params$threshold)
pct_above <- round(n_above / nrow(filtered_data) * 100, 1)
median_conc <- round(median(filtered_data$concentration), 2)
```

## Results from this parameter set

Of the `r nrow(filtered_data)` participants at **`r params$site`**, **`r n_above` (`r pct_above`%)** had serum `r params$chemical` concentrations exceeding the `r params$threshold` ng/mL threshold. The median concentration was `r median_conc` ng/mL.

```{r}
#| label: fig-distribution
#| fig-cap: !expr paste("Distribution of serum", params$chemical, "at", params$site, "—", params$study_year)
#| fig-width: 7
#| fig-height: 4
#| echo: false

ggplot(filtered_data, aes(x = concentration)) +
  geom_histogram(fill = "#0D9488", color = "white", bins = 25) +
  geom_vline(xintercept = params$threshold, color = "#DC2626",
             linewidth = 1, linetype = "dashed") +
  annotate("text", x = params$threshold + 0.3, y = Inf,
           label = paste("Threshold:", params$threshold, "ng/mL"),
           vjust = 2, hjust = 0, color = "#DC2626", size = 3.5) +
  labs(title = paste(params$chemical, "Serum Concentrations,",
                     params$site, params$study_year),
       x = paste(params$chemical, "(ng/mL)"),
       y = "Count") +
  theme_minimal()
```

::: {.callout-note}
## Dynamic captions
The `fig-cap` above uses `!expr` to evaluate an R expression: `!expr paste("Distribution of", params$chemical, "...")`. This is Quarto's way to compute caption text from parameters at render time. Without this, the caption would be a static string.
:::

## Re-rendering with different parameters

### From the R console

```{r}
#| eval: false

# Render for PFOS at Durham
quarto::quarto_render(
  "06_parameterized_reports.qmd",
  execute_params = list(
    chemical = "PFOS",
    threshold = 6.0,
    site = "Durham",
    study_year = 2024
  ),
  output_file = "report_PFOS_Durham_2024.html"
)
```

Each call produces a different rendered output from the same source.

### From the command line

```bash
quarto render 06_parameterized_reports.qmd -P chemical:PFOS -P threshold:6.0 -P site:Durham
```

The `-P` flag (or `--execute-param`) sets a parameter. Repeat for each one.

### The "Render with Parameters" button

In RStudio (and Posit Cloud's IDE), click the dropdown arrow next to the **Render** button and select **"Render with Parameters..."** — you get a dialog where you can fill in values interactively. Useful for one-off renders without writing code.

## Batch report generation

The real power: a loop that generates a report for every combination you care about.

```{r}
#| label: batch-demo
#| eval: false

# Generate a separate HTML report for each chemical-site combination
chemicals <- c("PFOA", "PFOS", "PFNA", "BPA")
sites <- c("Research Triangle Park", "Durham", "Chapel Hill")

# Conventional thresholds (μg/L or ng/mL)
thresholds <- list(PFOA = 4.0, PFOS = 6.0, PFNA = 2.0, BPA = 5.0)

for (chem in chemicals) {
  for (site in sites) {
    out_file <- paste0("report_", chem, "_",
                       gsub(" ", "_", site), "_2024.html")

    quarto::quarto_render(
      "06_parameterized_reports.qmd",
      execute_params = list(
        chemical = chem,
        site = site,
        threshold = thresholds[[chem]],
        study_year = 2024
      ),
      output_file = out_file
    )
  }
}
# Result: 12 HTML reports, one per chemical-site combination
```

Twelve reports from one source file. Add a new chemical or site, re-run the loop, you have new reports. The analysis logic stays in one place.

::: {.callout-tip}
## When this pattern shines
Parameterized reports earn their keep when:

- You have a **template** that should apply to many cases.
- The **analysis is identical**; only inputs change.
- Reports need to be **regenerated regularly** as new data arrives.

NIEHS-relevant cases: annual chemical-specific exposure summaries, multi-site monitoring reports, sub-population analyses, sensitivity analyses across multiple thresholds.
:::

## Parameter validation

Parameters from a user can be wrong. A robust template validates them at the top:

```{r}
#| eval: false

# At the top of your setup chunk
valid_chemicals <- c("PFOA", "PFOS", "PFNA", "PFHxS", "GenX", "BPA")
if (!params$chemical %in% valid_chemicals) {
  stop("Invalid chemical: ", params$chemical,
       ". Must be one of: ", paste(valid_chemicals, collapse = ", "))
}

if (!is.numeric(params$threshold) || params$threshold <= 0) {
  stop("Invalid threshold: ", params$threshold,
       ". Must be a positive number.")
}
```

A failed render with a clear error is better than a successful render of nonsense.

## R Markdown vs Quarto for parameters

Same syntax, almost identical behavior:

| Aspect | R Markdown | Quarto |
|--------|-----------|--------|
| YAML key | `params:` | `params:` (same) |
| Access in R | `params$x` | `params$x` (same) |
| Render with params (R) | `rmarkdown::render(..., params = list(...))` | `quarto::quarto_render(..., execute_params = list(...))` |
| Render with params (CLI) | N/A in core R Markdown | `quarto render ... -P key:value` |
| "Render with Parameters" button | Yes | Yes |
| Dynamic figure captions | Use `fig.cap = paste(...)` chunk option | `#\| fig-cap: !expr paste(...)` |

Parameterized reports are arguably **the** feature that justifies moving an existing R Markdown workflow to Quarto for any team doing recurring report generation — the CLI integration and YAML-style chunk options make batch generation more script-friendly.

## Exercises

### Exercise 6.1 — Try different parameter values

1. In this file's YAML, change `chemical: "PFOA"` to `chemical: "PFOS"`.
2. Change `threshold: 4.0` to `threshold: 6.0`.
3. Change `site: "Research Triangle Park"` to `site: "Durham"`.
4. Render. Observe how the title, prose, statistics, and plot all update.

### Exercise 6.2 — Render two versions from R

Open the R console and run:

```{r}
#| eval: false

quarto::quarto_render(
  "06_parameterized_reports.qmd",
  execute_params = list(chemical = "PFOA", site = "Durham"),
  output_file = "report_PFOA_Durham.html"
)

quarto::quarto_render(
  "06_parameterized_reports.qmd",
  execute_params = list(chemical = "PFOS", site = "Chapel Hill"),
  output_file = "report_PFOS_ChapelHill.html"
)
```

Open both HTML files. Confirm they have different titles, statistics, and plots.

## What's next

Quarto Module 7 — **Projects, Themes, and Publishing**: organize multi-document collections (websites, books), customize the look (themes, CSS), and publish to GitHub Pages, Quarto Pub, or Posit Connect Cloud.

Open [`07_projects_themes_publishing.qmd`](07_projects_themes_publishing.qmd).
